---
title: "disc_selection"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2026-01-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installations for first use:
```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("phyloseq")
# 
# if(!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Maaslin2")
# 
# install.packages("Matrix")
# install.packages("pgirmess")
# install.packages("agricolae")
# install.packages("ggpubr")
# install.packages("gt")
# install.packages("FSA")
```


#Load libraries and data:

Required libraries:
```{r}
library(vegan)
library (ggplot2)
library(ggpubr)
library(data.table)
library(cowplot)

library(readr)
library(phyloseq)
library(pgirmess)
library(agricolae)
library(readxl)
library(dplyr)
library(tidyr)
library(gt)
library(FSA)
library(Maaslin2)
# https://huttenhower.sph.harvard.edu/maaslin2/
library(tidyverse)
library(lmerTest)
library(phyloseq)
library(vegan)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)

```


```{r}
filname= "20260326_Overview samples for Animesh_addedclinicalparameters.xlsx"
#phy.red <- readRDS("phyloseq_discDC_red.rds")

phy_disc <- readRDS("phyloseq_discDC_red.rds")
phy_val1 <- readRDS("phyloseq_val_one.rds")
phy_val2 <- readRDS("phyloseq_val_two.rds")

meta_phy <- data.frame(sample_data(phy_disc)) %>%
  mutate(Sample_ID = rownames(.)) %>%
  rename(Patient_ID = Patient.ID)

meta <- read_excel(filname, sheet = "Metadata") %>%
  rename(
    Patient_ID = `Patient-ID`,
    Gestational_age = `Gestational age (days)`,
    Birth_weight = `Birth weight (grams)`,
    Center = `Center of birth`,
    AB_ratio = AB_IV_days_beforet0_ratio,
    Feeding_type = `Feeding type at full enteral feeding`,
    Mode_delivery = `Mode of delivery`,
    Week_of_life = Sample_week_of_life,
    Invasive_device = Invasive_medical_device_beforet0,
    Storage_years = Storage_duration_years
  )

meta_clean <- meta %>%
  group_by(Patient_ID) %>%
  summarise(across(everything(), ~ first(na.omit(.))), .groups = "drop")

df <- meta_phy %>%
  select(-any_of("Cohort")) %>%   # remove duplicate if exists
  left_join(meta_clean, by = "Patient_ID") %>%
  filter(!is.na(Cohort))



prepare_phyloseq <- function(phy_obj, cohort_name, meta_clean) {
  
  # Extract phyloseq metadata
  meta_phy <- data.frame(sample_data(phy_obj))
  
  # 🔥 ALWAYS define Sample_ID from phyloseq
  meta_phy$Sample_ID <- phyloseq::sample_names(phy_obj)
  
  # Rename Patient ID (adjust if needed)
  meta_phy <- meta_phy %>%
    rename(Patient_ID = Patient.ID)
  
  # Ensure same type
  meta_phy$Patient_ID <- as.character(meta_phy$Patient_ID)
  meta_clean$Patient_ID <- as.character(meta_clean$Patient_ID)
  
  # 🔥 Merge
  df <- left_join(meta_phy, meta_clean, by = "Patient_ID")
  
  # 🔥 FORCE Sample_ID again (critical fix)
  df$Sample_ID <- phyloseq::sample_names(phy_obj)
  
  # Align samples
  common_samples <- intersect(sample_names(phy_obj), df$Sample_ID)
  
  if(length(common_samples) == 0){
    stop(paste("❌ No matching samples in", cohort_name))
  }
  
  # Subset phyloseq
  phy_obj <- prune_samples(common_samples, phy_obj)
  
  # Subset df
  df <- df %>%
    filter(Sample_ID %in% common_samples) %>%
    distinct(Sample_ID, .keep_all = TRUE)
  
  # 🔥 CRITICAL: correct rownames
  rownames(df) <- df$Sample_ID
  
  sample_data(phy_obj) <- sample_data(df)
  
  df$Cohort <- cohort_name
  
  return(list(phy = phy_obj, df = df))
}

disc <- prepare_phyloseq(phy_disc, "Discovery cohort", meta_clean)
val1 <- prepare_phyloseq(phy_val1, "Validation cohort 1", meta_clean)
val2 <- prepare_phyloseq(phy_val2, "Validation cohort 2", meta_clean)


```



```{r}

clinical_vars <- c(
  "Disease_state",
  "Subgroup",
  "Bloodculture_isolated_pathogen.x",
  "Gestational_age",
  "Birth_weight",
  "Center",
  "AB_ratio",
  "Feeding_type",
  "Mode_delivery",
  "Week_of_life",
  "Invasive_device",
  "Storage_years"
)

run_permanova <- function(phy_obj, df, var) {
  
  df_sub <- df %>%
    select(Sample_ID, Patient_ID, all_of(var)) %>%
    drop_na()
  
  if(nrow(df_sub) < 5) return(NULL)
  if(length(unique(df_sub[[var]])) < 2) return(NULL)
  
  dist_mat <- phyloseq::distance(phy_obj, method = "bray")
  
  dist_sub <- as.dist(as.matrix(dist_mat)[df_sub$Sample_ID, df_sub$Sample_ID])
  
  model <- vegan::adonis2(
    dist_sub ~ df_sub[[var]],
    permutations = 999,
    strata = df_sub$Patient_ID
  )
  
  data.frame(
    Variable = var,
    R2 = model$R2[1],
    p = model$`Pr(>F)`[1]
  )
}

library(purrr)

run_cohort <- function(obj, cohort_name) {
  
  phy_obj <- obj$phy
  df <- obj$df
  
  res <- map_df(clinical_vars, function(v) {
    out <- run_permanova(phy_obj, df, v)
    if(is.null(out)) return(NULL)
    out
  })
  
  res$Cohort <- cohort_name
  return(res)
}

res_disc <- run_cohort(disc, "Discovery cohort")
res_val1 <- run_cohort(val1, "Validation cohort 1")
res_val2 <- run_cohort(val2, "Validation cohort 2")

results_beta <- bind_rows(res_disc, res_val1, res_val2)

```



```{r}
results_beta <- results_beta %>%
  group_by(Cohort) %>%
  mutate(p_adj = p.adjust(p, method = "fdr")) %>%
  ungroup()

results_beta <- results_beta %>%
  mutate(
    Variable_clean = case_when(
      grepl("Gestational_age", Variable) ~ "Gestational age",
      grepl("Week_of_life", Variable) ~ "Week of life",
      grepl("Birth_weight", Variable) ~ "Birth weight",
      grepl("Center", Variable) ~ "Center of birth",
      grepl("AB_ratio", Variable) ~ "Antibiotic exposure",
      grepl("Feeding_type", Variable) ~ "Feeding type",
      grepl("Mode_delivery", Variable) ~ "Mode of delivery",
      grepl("Invasive_device", Variable) ~ "Invasive device",
      grepl("Storage_years", Variable) ~ "Storage duration",
      TRUE ~ Variable
    )
  )

library(ggplot2)

ggplot(results_beta,
       aes(x = Cohort,
           y = reorder(Variable_clean, R2),
           fill = R2)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = ifelse(p_adj < 0.05, "*", "")),
    size = 5
  ) +
  
  scale_fill_gradient(low = "white", high = "red") +
  
  labs(
    title = "Clinical drivers of beta-diversity (PERMANOVA)",
    x = "Cohort",
    y = "Clinical parameters",
    fill = expression(R^2)
  ) +
  
  theme_minimal(base_size = 12)
```



```{r}
# Or Excel
library(writexl)
write_xlsx(results_beta,
           "rawfileAB/beta_diversity_permanova_results_all_cohorts.xlsx")

write.csv(res_disc,
          "rawfileAB/permanova_discovery_cohort.csv",
          row.names = FALSE)

write.csv(res_val1,
          "rawfileAB/permanova_validation1_cohort.csv",
          row.names = FALSE)

write.csv(res_val2,
          "rawfileAB/permanova_validation2_cohort.csv",
          row.names = FALSE)

results_clean <- results_beta %>%
  select(Cohort,
         Variable = Variable_clean,
         R2,
         p_value = p,
         FDR = p_adj) %>%
  arrange(Cohort, desc(R2))

write_xlsx(results_clean,
           "rawfileAB/PERMANOVA_clean_results.xlsx")
```







```{r}
summary(results_beta$p_adj)
```



```{r}

results_beta <- results_beta %>%
  mutate(across(c(Variable, Variable_clean), ~ gsub("\\.x$", "", .)))

results_beta <- results_beta %>%
  mutate(
    signif_label = case_when(
      R2 >= 0.20 ~ "***",
      R2 >= 0.12 ~ "**",
      R2 >= 0.02 ~ "*",
      TRUE ~ ""
    )
  )


ggplot(results_beta,
       aes(x = Cohort,
           y = reorder(Variable_clean, R2),
           fill = R2)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = signif_label),
    size = 6,
    color = "black"
  ) +
  
  scale_fill_gradient(low = "white", high = "red") +
  
  labs(
    title = "Clinical drivers of beta-diversity (PERMANOVA)",
    x = "Cohort",
    y = "Clinical parameters",
    fill = expression(R^2)
  ) +
  
  theme_minimal(base_size = 12)

##

```

## Stars represent effect size based on R² thresholds (* ≥ 0.05, ** ≥ 0.10, *** ≥ 0.15), not statistical significance.

```{r}
p <- ggplot(results_beta,
       aes(x = Cohort,
           y = reorder(Variable_clean, R2),
           fill = R2)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = signif_label),
    size = 6,
    color = "black"
  ) +
  
  scale_fill_gradient(low = "white", high = "red") +
  
  labs(
    title = "Clinical drivers of beta-diversity (PERMANOVA)",
    x = "Cohort",
    y = "Clinical parameters",
    fill = expression(R^2)
  ) +
  
  theme_minimal(base_size = 12)

ggsave(
  filename = "figs/BetaDiversity_PERMANOVA_Heatmap.png",
  plot = p,
  width = 10,
  height = 8,
  dpi = 300
)

ggsave(
  filename = "figs/BetaDiversity_PERMANOVA_Heatmap.pdf",
  plot = p,
  width = 10,
  height = 8
)

```


### till now ok










```{r}

filname= "20260326_Overview samples for Animesh_addedclinicalparameters.xlsx"
#phy.red <- readRDS("phyloseq_discDC_red.rds")

phy_disc <- readRDS("phyloseq_discDC_red.rds")
phy_val1 <- readRDS("phyloseq_val_one.rds")
phy_val2 <- readRDS("phyloseq_val_two.rds")

meta_phy <- data.frame(sample_data(phy_disc)) %>%
  mutate(Sample_ID = rownames(.)) %>%
  rename(Patient_ID = Patient.ID)

meta <- read_excel(filname, sheet = "Metadata") %>%
  rename(
    Patient_ID = `Patient-ID`,
    Gestational_age = `Gestational age (days)`,
    Birth_weight = `Birth weight (grams)`,
    Center = `Center of birth`,
    AB_ratio = AB_IV_days_beforet0_ratio,
    Feeding_type = `Feeding type at full enteral feeding`,
    Mode_delivery = `Mode of delivery`,
    Week_of_life = Sample_week_of_life,
    Invasive_device = Invasive_medical_device_beforet0,
    Storage_years = Storage_duration_years
  )

meta_clean <- meta %>%
  group_by(Patient_ID) %>%
  summarise(across(everything(), ~ first(na.omit(.))), .groups = "drop")

df <- meta_phy %>%
  select(-any_of("Cohort")) %>%   # remove duplicate if exists
  left_join(meta_clean, by = "Patient_ID") %>%
  filter(!is.na(Cohort))



## may need code here 

library(phyloseq)

dist_mat <- phyloseq::distance(phy.red, method = "bray")

clinical_vars <- c(
  "Disease_state",
  "Subgroup",
  "Bloodculture_isolated_pathogen",
  "Gestational_age",
  "Birth_weight",
  "Center",
  "AB_ratio",
  "Feeding_type",
  "Mode_delivery",
  "Invasive_device",
  "Storage_years"
)


```

```{r}

colnames(df)
```



```{r}

library(vegan)
library(dplyr)

run_permanova <- function(data, dist_mat, var){
  
  df_sub <- data %>%
    select(Sample_ID, Patient_ID, all_of(var)) %>%
    drop_na()
  
  if(nrow(df_sub) < 5) return(NULL)
  
  if(length(unique(df_sub[[var]])) < 2) return(NULL)
  
  # Match distance matrix
  dist_sub <- as.dist(as.matrix(dist_mat)[df_sub$Sample_ID, df_sub$Sample_ID])
  
  formula <- as.formula(paste("dist_sub ~", var))
  
  model <- adonis2(
    formula,
    data = df_sub,
    permutations = 999,
    strata = df_sub$Patient_ID   # 🔥 repeated measures correction
  )
  
  data.frame(
    Variable = var,
    R2 = model$R2[1],
    p = model$`Pr(>F)`[1]
  )
}

library(dplyr)
library(purrr)

results_beta <- df %>%
  group_split(Cohort) %>%
  purrr::map_df(function(subdf) {
    
    cohort_name <- unique(subdf$Cohort)
    
    res <- purrr::map_df(clinical_vars, function(v) {
      out <- run_permanova(subdf, v)
      if(is.null(out)) return(NULL)
      out
    })
    
    res$Cohort <- cohort_name
    res
  })

results_beta <- results_beta %>%
  group_by(Cohort) %>%
  mutate(p_adj = p.adjust(p, method = "fdr")) %>%
  ungroup()

results_beta <- results_beta %>%
  mutate(
    Variable_clean = case_when(
      grepl("Gestational_age", Variable) ~ "Gestational age",
      grepl("Birth_weight", Variable) ~ "Birth weight",
      grepl("Center", Variable) ~ "Center of birth",
      grepl("AB_ratio", Variable) ~ "Antibiotic exposure",
      grepl("Feeding_type", Variable) ~ "Feeding type",
      grepl("Mode_delivery", Variable) ~ "Mode of delivery",
      grepl("Invasive_device", Variable) ~ "Invasive device",
      grepl("Storage_years", Variable) ~ "Storage duration",
      TRUE ~ Variable
    )
  )

library(ggplot2)

ggplot(results_beta,
       aes(x = Cohort,
           y = reorder(Variable_clean, R2),
           fill = R2)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = ifelse(p_adj < 0.05, "*", "")),
    size = 5
  ) +
  
  scale_fill_gradient(low = "white", high = "red") +
  
  labs(
    title = "Clinical drivers of beta-diversity (PERMANOVA)",
    x = "Cohort",
    y = "Clinical parameters",
    fill = expression(R^2)
  ) +
  
  theme_minimal(base_size = 12)

```


```{r}
results_clean <- results %>%
  filter(term != "(Intercept)") %>%
  filter(!grepl("Disease_state", term)) %>%
  mutate(
    p_adj = p.adjust(p.value, method = "fdr"),
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

results_clean <- results_clean %>%
  mutate(
    term_clean = case_when(
      grepl("Bloodculture", term) ~ 
        gsub("Bloodculture_isolated_pathogen", "", term),
      grepl("Gestational_age", term) ~ "Gestational age",
      grepl("Birth_weight", term) ~ "Birth weight",
      grepl("Center", term) ~ "Center of birth",
      grepl("AB_ratio", term) ~ "Antibiotic exposure",
      grepl("Feeding_type", term) ~ "Feeding type",
      grepl("Mode_delivery", term) ~ "Mode of delivery",
      grepl("Invasive_device", term) ~ "Invasive device",
      grepl("Storage_years", term) ~ "Storage duration",
      TRUE ~ term
    )
  )

ggplot(results_clean,
       aes(x = estimate,
           y = reorder(term_clean, estimate))) +
  
  geom_point(size = 3) +
  
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  
  geom_vline(xintercept = 0, linetype = "dashed") +
  
  geom_text(
    aes(label = significance),
    hjust = -0.3,
    size = 4
  ) +
  
  facet_grid(Alpha ~ Cohort, scales = "free_y") +
  
  labs(
    x = "Effect size (β ± 95% CI)",
    y = "Clinical parameters"
  ) +
  
  theme_bw(base_size = 12)
```




```{r}


results_clean <- results_clean %>%
  mutate(
    term_clean = case_when(
      
      grepl("Bloodculture_isolated_pathogen", term) ~ 
        gsub("Bloodculture_isolated_pathogen\\.x", "", term) %>%
        gsub("Bloodculture_isolated_pathogen\\.y", "", .),
      
      grepl("Gestational_age", term) ~ "Gestational age",
      grepl("Birth_weight", term) ~ "Birth weight",
      grepl("Center", term) ~ "Center of birth",
      grepl("AB_ratio", term) ~ "Antibiotic exposure",
      grepl("Feeding_type", term) ~ "Feeding type",
      grepl("Mode_delivery", term) ~ "Mode of delivery",
      grepl("Invasive_device", term) ~ "Invasive device",
      grepl("Storage_years", term) ~ "Storage duration",
      
      TRUE ~ term
    )
  )

library(stringr)

results_clean <- results_clean %>%
  mutate(term_clean = str_replace_all(term_clean, "^\\.", ""))  # remove leading dots


plot_alpha <- function(alpha_name) {
  
  p <- results_clean %>%
    filter(Alpha == alpha_name) %>%
    ggplot(aes(x = estimate,
               y = reorder(term_clean, estimate))) +
    
    geom_point(size = 3) +
    
    geom_errorbarh(
      aes(xmin = conf.low, xmax = conf.high),
      height = 0.2
    ) +
    
    geom_vline(xintercept = 0, linetype = "dashed") +
    
    geom_text(
      aes(label = significance),
      hjust = -0.3,
      size = 4
    ) +
    
    facet_wrap(~Cohort, ncol = 1, scales = "free_y") +
    
    labs(
      title = paste("Alpha Diversity:", alpha_name),
      x = "Effect size (β ± 95% CI)",
      y = "Clinical parameters"
    ) +
    
    theme_bw(base_size = 12)
  
  return(p)
}

p_shannon  <- plot_alpha("Shannon")
p_simpson  <- plot_alpha("Simpson")
p_observed <- plot_alpha("Observed")

ggsave("figs/fig1-Shannon_effects.png", p_shannon, width = 8, height = 10)
ggsave("figs/fig1-Simpson_effects.png", p_simpson, width = 8, height = 10)
ggsave("figs/fig1-Observed_effects.png", p_observed, width = 8, height = 10)

```




```{r}
plot_alpha <- function(alpha_name) {
  
  p <- results_clean %>%
    filter(Alpha == alpha_name) %>%
    ggplot(aes(x = estimate,
               y = reorder(term_clean, estimate))) +
    
    geom_point(size = 3) +
    
    geom_errorbarh(
      aes(xmin = conf.low, xmax = conf.high),
      height = 0.2
    ) +
    
    geom_vline(xintercept = 0, linetype = "dashed") +
    
    geom_text(
      aes(label = significance),
      hjust = -0.3,
      size = 4
    ) +
    
    facet_wrap(~Cohort, ncol = 1, scales = "free_y") +
    
    labs(
      title = paste("Alpha Diversity:", alpha_name),
      x = "Effect size (β ± 95% CI)",
      y = "Clinical parameters"
    ) +
    
    theme_bw(base_size = 12)
  
  return(p)
}

p_shannon  <- plot_alpha("Shannon")
p_simpson  <- plot_alpha("Simpson")
p_observed <- plot_alpha("Observed")

```


```{r}
names(results_clean)
```



```{r}

table(df$Cohort)

```

How to interpret your figure (for paper)
β > 0 → increases microbiome diversity
β < 0 → decreases diversity
CI crossing 0 → not significant
Compare consistency across:
Cohort A (discovery)
Cohort B/C (validation)


```{r}
ggsave("Figure1_AlphaDiversity_ClinicalFactors.png",
       width = 10, height = 6, dpi = 300)


write.csv(results_clean,
          "Figure1_LMM_results.csv",
          row.names = FALSE)
```







