
# 

#Load libraries and data:

Required libraries:
```{r}
library(vegan)
library (ggplot2)
library(data.table)
library(cowplot)
#library(Maaslin2)
library(readr)
library(phyloseq)
library(pgirmess)
library(agricolae)
library(readxl) 
library(caret)
library(tidyverse)
library(microbiome)
library(MLmetrics)

```

#How to make phyloseq object:

Reformat table:


```{r}
# ------------------------------------------------------------
# 2️⃣ Read OTU table (Discovery cohort - DC)
# ------------------------------------------------------------
discDC <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                     sheet = "DC") %>% as.data.frame()

# First column contains sample IDs → make them rownames
rownames(discDC) <- discDC[[1]]
discDC <- discDC[ , -1]

# Remove unwanted PC samples (rows)
remove_samples <- paste0("DC-PC", 1:13)
discDC <- discDC[!rownames(discDC) %in% remove_samples, ]
cat("Number of samples after removing PC samples:", nrow(discDC), "\n")

# Transpose: rows = OTUs, cols = samples
DCotumat <- t(discDC)

# ------------------------------------------------------------
# 3️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# Keep only OTUs present in both OTU and taxonomy tables
common_otus <- intersect(rownames(DCotumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# Subset OTU and taxonomy tables
DCotumat_common <- DCotumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# Sanity check
dim(DCotumat_common)
dim(taxonomy_common)

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(DCotumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqDC <- phyloseq(OTU, TAX)
cat("physeqDC - Samples:", nsamples(physeqDC), "OTUs:", ntaxa(physeqDC), "\n")

## new added
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Discovery cohort", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"

# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(DCotumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqDC <- merge_phyloseq(physeqDC, sampledata)

cat("physeqDC with metadata - Samples:", nsamples(physeqDC), 
    "OTUs:", ntaxa(physeqDC), "\n")


```


# If you also have SV1 and SV2 phyloseq objects

# prep SV1
```{r}

# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV1
# ------------------------------------------------------------
discSV1 <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx", 
                      sheet = "SV 1") %>% as.data.frame()

# first column = sample IDs
rownames(discSV1) <- discSV1[[1]]
discSV1 <- discSV1[ , -1]

# transpose: rows = OTUs, cols = samples
SV1otumat <- t(discSV1)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV1otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# subset OTU and taxonomy tables
SV1otumat_common <- SV1otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV1otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV1 <- phyloseq(OTU, TAX)

cat("physeqSV1 - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 1", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"

# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV1otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV1 <- merge_phyloseq(physeqSV1, sampledata)

cat("physeqSV1 with metadata - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

```

# prep SV2 data
```{r}
# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV2
# ------------------------------------------------------------
discSV2 <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx", 
                      sheet = "SV 2") %>% as.data.frame()

# Suppose the first column contains the row IDs (sample names)
rownames(discSV2) <- discSV2[[1]]
discSV2 <- discSV2[ , -1]  # remove the first column after assigning rownames

# Now remove rows that contain "SV2-Blanco"
discSV2 <- discSV2[!grepl("^SV2-Blanco", rownames(discSV2)), ]

cat("Number of rows after removing blanks:", nrow(discSV2), "\n")

# transpose: rows = OTUs, cols = samples
SV2otumat <- t(discSV2)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV2otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

SV2otumat_common <- SV2otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV2otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV2 <- phyloseq(OTU, TAX)

cat("physeqSV2 - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 2
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 2", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"

# Keep only rows that exist in SV2 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV2otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV2 <- merge_phyloseq(physeqSV2, sampledata)

cat("physeqSV2 with metadata - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

```

# til sv2

## newruns for species level

```{r}

# Assume physeqDC already has sample_data merged
physeq_full <- physeqDC   # no time-period filtering
# Collapse OTUs to species level
physeq_full_species <- tax_glom(physeq_full, taxrank = "Species") # Species
physeqSV1_species    <- tax_glom(physeqSV1, taxrank = "Species")
physeqSV2_species    <- tax_glom(physeqSV2, taxrank = "Species")

# If Species has many missing values, you could use Genus instead:
  # tax_glom(physeq_full, taxrank = "Genus")

```


```{r}
# Compute common features across cohorts

# Extract OTU tables (now species-level)
otu_full    <- as.data.frame(t(otu_table(physeq_full_species)))
otu_sv1     <- as.data.frame(t(otu_table(physeqSV1_species)))
otu_sv2     <- as.data.frame(t(otu_table(physeqSV2_species)))

# Keep only species present in all three cohorts
common_species <- Reduce(intersect, list(
  colnames(otu_full),
  colnames(otu_sv1),
  colnames(otu_sv2)
))
cat("Number of common species:", length(common_species), "\n")


```


# Subset OTU tables to common species CLR transform and z-normalize


```{r}

otu_full <- otu_full[, common_species, drop = FALSE]
otu_sv1  <- otu_sv1[, common_species, drop = FALSE]
otu_sv2  <- otu_sv2[, common_species, drop = FALSE]

# transform and z-normalize
# Add 1 (or 0.5) to all counts to avoid zeros
otu_full_pseudo <- otu_full + 1
otu_full_clr   <- microbiome::transform(otu_full_pseudo, "clr")
#otu_full_clr   <- microbiome::transform(otu_full, "clr")
otu_full_clr_z <- scale(otu_full_clr)
train_mean     <- attr(otu_full_clr_z, "scaled:center")
train_sd       <- attr(otu_full_clr_z, "scaled:scale")

# Combine with metadata
meta_full <- data.frame(sample_data(physeq_full_species))
meta_full$DiseaseStatus <- ifelse(
  meta_full$DiseaseStatus %in% c("LOS","Gut-derived LOS","1"),
  "LOS","Control"
)
meta_full$DiseaseStatus <- factor(meta_full$DiseaseStatus, levels = c("Control","LOS"))

#dc_df <- as.data.frame(otu_full_clr_z) %>%
#         mutate(DiseaseStatus = meta_full$DiseaseStatus)

# CLR + z-normalized OTU/species table
dc_df <- as.data.frame(otu_full_clr_z)
# 2️⃣ Keep only numeric columns
num_cols <- sapply(dc_df, is.numeric)
dc_df_num <- dc_df[, num_cols, drop = FALSE]
# 3️⃣ Replace NaN/Inf with 0
dc_df_num[!is.finite(as.matrix(dc_df_num))] <- 0
# 4️⃣ Remove zero-variance columns
nonzero_var <- apply(dc_df_num, 2, function(x) var(x, na.rm = TRUE) > 0)
# Check if we have any columns left
if(sum(nonzero_var) == 0){
  stop("No non-zero-variance features available after cleaning.")
}

dc_df_num <- dc_df_num[, nonzero_var, drop = FALSE]
# 5️⃣ Add DiseaseStatus
dc_df <- dc_df_num
dc_df$DiseaseStatus <- meta_full$DiseaseStatus
# 6️⃣ Quick check
cat("Number of predictors:", ncol(dc_df) - 1, "\n")
table(dc_df$DiseaseStatus)

```


## runs- show DC metrics

```{r}
# Impute missing values in the full dataframe
dc_df[is.na(dc_df)] <- 0   # replace all NA with 0
## --- Custom summary function ----------------------------------
train_index <- createDataPartition(dc_df$DiseaseStatus, p = 0.8, list = FALSE)
train_set   <- dc_df[train_index, ]
test_set    <- dc_df[-train_index, ]

multiSummary <- function(data, lev = NULL, model = NULL) {
  # data has: obs (true labels), pred (predicted labels), prob for each class
  out <- c()
  
  # Sensitivity & Specificity
  sens <- sensitivity(data$pred, data$obs, positive = lev[2])
  spec <- specificity(data$pred, data$obs, negative = lev[1])
  
  # AUROC
  roc  <- tryCatch(pROC::roc(response = data$obs,
                             predictor = data[, lev[2]],
                             levels = rev(lev),
                             quiet = TRUE)$auc,
                   error = function(e) NA)
  
  # Precision, Recall, F1
  prec <- Precision(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  rec  <- Recall(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  f1   <- F1_Score(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  
  out <- c(ROC = roc, Sens = sens, Spec = spec,
           Precision = prec, Recall = rec, F1 = f1)
  return(out)
}

## --- Train Control --------------------------------------------
ctrl <- trainControl(
  method = "repeatedcv",
  number = 10, repeats = 3,
  classProbs = TRUE,
  summaryFunction = multiSummary,
  savePredictions = "final"
)

## --- Train model ----------------------------------------------
set.seed(123)
rf_model <- train(
  DiseaseStatus ~ .,
  data       = train_set,
  method     = "ranger",
  metric     = "ROC",
  tuneGrid   = expand.grid(
                 mtry          = floor(sqrt(ncol(train_set) - 1)),
                 splitrule     = "gini",
                 min.node.size = 1
               ),
  trControl  = ctrl,
  importance = "impurity"
)

## --- Extract results ------------------------------------------
results <- rf_model$resample[, c("ROC","Sens","Spec","Precision","Recall","F1")]

summary_table <- data.frame(
  Average = colMeans(results, na.rm = TRUE),
  SD      = apply(results, 2, sd, na.rm = TRUE)
)

# transpose so rows = Average/SD, columns = metrics
summary_table <- t(summary_table)

print(summary_table)



```

# extra for ROC
```{r}
## ------------------------------------------------------------
## Libraries
## ------------------------------------------------------------
library(pROC)
library(ggplot2)
library(dplyr)

## ------------------------------------------------------------
## Extract cross-validated predictions from caret model
## ------------------------------------------------------------
pred <- rf_model$pred

# Ensure outcome is a factor with correct levels
pred$obs <- factor(pred$obs)

# Identify positive class (caret convention: 2nd level)
positive_class <- levels(pred$obs)[2]

## ------------------------------------------------------------
## Compute ROC and AUROC
## ------------------------------------------------------------
roc_discovery <- roc(
  response  = pred$obs,
  predictor = pred[[positive_class]],
  levels    = levels(pred$obs),
  direction = "<"
)

auc_val <- auc(roc_discovery)

###

roc_plot <- ggroc(
  roc_discovery,
  colour = "black",
  size = 1.2,
  legacy.axes = TRUE
) +
  ggtitle(paste0(
    "ROC Curve – Discovery Cohort (CV)\nAUC = ",
    round(auc_val, 3)
  )) +
  theme_minimal() +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey50"
  ) +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

# Show plot
print(roc_plot)

# Save as PDF
ggsave(
  filename = "6aROC_Curve_Discovery.pdf",
  plot = roc_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)



```


## PR-AUC for Discovery Cohort (CV)

```{r}
library(PRROC)
pred <- rf_model$pred

# Ensure factor
pred$obs <- factor(pred$obs)

# Positive class (Ecoli_LOS)
positive_class <- levels(pred$obs)[2]

# Convert to binary labels
truth <- ifelse(pred$obs == positive_class, 1, 0)

# Predicted probabilities
scores <- pred[[positive_class]]
pr_discovery <- pr.curve(
  scores.class0 = scores[truth == 1],
  scores.class1 = scores[truth == 0],
  curve = TRUE
)

pr_auc_val <- pr_discovery$auc.integral

print(paste("Discovery PR-AUC:", round(pr_auc_val, 3)))
baseline <- mean(truth)
plot(
  pr_discovery,
  auc.main = FALSE,
  main = paste0(
    "PR Curve – Discovery Cohort (CV)\nPR-AUC = ",
    round(pr_auc_val, 3),
    " | baseline = ",
    round(baseline, 3)
  )
)

abline(h = baseline, col = "red", lty = 2)

pdf("PR_Curve_Discovery.pdf", width = 8, height = 6)

plot(
  pr_discovery,
  auc.main = FALSE,
  main = paste0(
    "PR Curve – Discovery Cohort (CV)\nPR-AUC = ",
    round(pr_auc_val, 3),
    " | baseline = ",
    round(baseline, 3)
  )
)

abline(h = baseline, col = "red", lty = 2)

dev.off()
```





```{r}

library(ggplot2)

# Extract variable importance
var_imp <- varImp(rf_model)$importance

# Add Original_ID column directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending
var_imp_ordered <- var_imp[order(-var_imp$Overall), ]

# Select top 20
top20 <- head(var_imp_ordered, 20)

# Plot
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()


```




# new ml run for top features
```{r}
library(caret)
library(ranger)
library(pROC)
library(MLmetrics)
library(dplyr)

# -------------------------------
# 1️⃣ Extract top significant features from the first RF model
# -------------------------------

# Get variable importance
rf_varimp <- varImp(rf_model, scale = TRUE)$importance

# Sort by importance (descending) and take top N features
top_n <- 10  # choose how many top features you want
sig_features <- rownames(rf_varimp)[order(rf_varimp$Overall, decreasing = TRUE)][1:top_n]

# Add the outcome column
sig_features <- c(sig_features, "DiseaseStatus")

# Subset the dataframe to only these features
dc_sig_df <- dc_df[, sig_features]

# -------------------------------
# 2️⃣ Train-test split
# -------------------------------
set.seed(123)
train_index <- createDataPartition(dc_sig_df$DiseaseStatus, p = 0.8, list = FALSE)
train_set <- dc_sig_df[train_index, ]
test_set  <- dc_sig_df[-train_index, ]

# -------------------------------
# 3️⃣ Custom summary function
# -------------------------------
multiSummary <- function(data, lev = NULL, model = NULL) {
  out <- c()
  sens <- sensitivity(data$pred, data$obs, positive = lev[2])
  spec <- specificity(data$pred, data$obs, negative = lev[1])
  roc  <- tryCatch(
    pROC::roc(response = data$obs, predictor = data[, lev[2]], levels = rev(lev), quiet = TRUE)$auc,
    error = function(e) NA
  )
  prec <- Precision(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  rec  <- Recall(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  f1   <- F1_Score(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  out <- c(ROC = roc, Sens = sens, Spec = spec, Precision = prec, Recall = rec, F1 = f1)
  return(out)
}

# -------------------------------
# 4️⃣ Train control
# -------------------------------
ctrl <- trainControl(
  method = "repeatedcv",
  number = 10, repeats = 3,
  classProbs = TRUE,
  summaryFunction = multiSummary,
  savePredictions = "final"
)

# -------------------------------
# 5️⃣ Train RF on top features
# -------------------------------
set.seed(123)
rf_top_model <- train(
  DiseaseStatus ~ .,
  data       = train_set,
  method     = "ranger",
  metric     = "ROC",
  tuneGrid   = expand.grid(
    mtry          = floor(sqrt(ncol(train_set) - 1)),
    splitrule     = "gini",
    min.node.size = 1
  ),
  trControl  = ctrl,
  importance = "impurity"
)

# -------------------------------
# 6️⃣ Extract CV results
# -------------------------------
results_top <- rf_top_model$resample[, c("ROC","Sens","Spec","Precision","Recall","F1")]

summary_table_top <- data.frame(
  Average = colMeans(results_top, na.rm = TRUE),
  SD      = apply(results_top, 2, sd, na.rm = TRUE)
)

# Transpose for readability
summary_table_top <- t(summary_table_top)

print(summary_table_top)

# -------------------------------
# 7️⃣ Optional: plot variable importance for top features
# -------------------------------
varImp_top <- varImp(rf_top_model, scale = TRUE)
print(varImp_top)


```




# change every time file name 
```{r}

library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
library(readxl)

# --- 1️⃣ Extract variable importance ---
var_imp <- varImp(rf_model)$importance

# Add Original_ID directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending and add Rank
var_imp <- var_imp[order(-var_imp$Overall), ]
var_imp$Rank <- 1:nrow(var_imp)

# Top 20
top20 <- head(var_imp, 20)

# --- 2️⃣ Plot top 20 OTUs ---
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()

# --- 3️⃣ Merge taxonomy info ---
tax <- read_excel(
  "20250512_OTU tables_percohort_sepsis prediction model.xlsx",
  sheet = "OTU table all"
) %>%
  mutate(ID.1 = as.character(ID.1))

top20 <- top20 %>%
  mutate(Original_ID = as.character(Original_ID)) %>%
  left_join(tax %>% select(ID.1, Family, Genus, Species),
            by = c("Original_ID" = "ID.1")) %>%
  select(Rank, Original_ID, Family, Genus, Species, Overall)

# --- 4️⃣ Create table grob ---
tab_grob <- tableGrob(
  top20,
  rows = NULL,
  theme = ttheme_default(
    core   = list(fg_params = list(cex = 0.8)),
    colhead= list(fg_params = list(cex = 1, fontface = "bold"))
  )
)

title <- textGrob("Top 20 OTUs with Taxonomy",
                  gp = gpar(fontsize = 16, fontface = "bold"))

# --- 5️⃣ Save as PNG ---
png("Species_Top20_OTU_Taxonomy.png", width = 2000, height = 1400, res = 200)
grid.arrange(title, tab_grob, ncol = 1, heights = c(0.1, 0.8, 0.1))
dev.off()



```



# --- Function to prepare validation set ---

```{r}
prepare_validation <- function(physeq_obj, train_mean, train_sd, dc_features) {
  
  # 1️⃣ Extract OTU table (samples × features)
  otu_df <- as.data.frame(t(otu_table(physeq_obj)))
  otu_df[] <- lapply(otu_df, as.numeric)

  # 2️⃣ Keep shared features
  otu_df <- otu_df[, intersect(colnames(otu_df), dc_features), drop = FALSE]

  # 3️⃣ Add missing features as zeros (SAFE)
  missing_feats <- setdiff(dc_features, colnames(otu_df))
  if (length(missing_feats) > 0) {
    zero_mat <- matrix(
      0,
      nrow = nrow(otu_df),
      ncol = length(missing_feats),
      dimnames = list(rownames(otu_df), missing_feats)
    )
    otu_df <- cbind(otu_df, zero_mat)
  }

  # 4️⃣ Reorder to discovery feature order
  otu_df <- otu_df[, dc_features, drop = FALSE]

  # 5️⃣ Add pseudocount (CRITICAL)
  otu_df <- otu_df + 1e-6

  # 6️⃣ CLR transform (MUST be matrix)
  otu_clr <- microbiome::transform(as.matrix(otu_df), "clr")

  # 7️⃣ Z-score using discovery stats
  otu_clr_z <- scale(otu_clr, center = train_mean, scale = train_sd)

  # 8️⃣ Metadata + label
  meta <- data.frame(sample_data(physeq_obj))
  meta$DiseaseStatus <- ifelse(
    meta$DiseaseStatus %in% c("LOS", "Gut-derived LOS", "1"),
    "LOS",
    "Control"
  )
  meta$DiseaseStatus <- factor(meta$DiseaseStatus, levels = c("Control", "LOS"))

  # 9️⃣ Combine
  df <- as.data.frame(otu_clr_z)
  df$DiseaseStatus <- meta$DiseaseStatus

  return(df)
}



# --- Prepare SV1 validation set ---
sv1_df <- prepare_validation(physeqSV1, train_mean, train_sd, colnames(otu_full))

# --- Prepare SV2 validation set ---
sv2_df <- prepare_validation(physeqSV2, train_mean, train_sd, colnames(otu_full))

nrow(sv1_df); nrow(sv2_df)

remove_zero_var <- function(df, label = "DiseaseStatus") {

  stopifnot(label %in% colnames(df))

  feats <- setdiff(colnames(df), label)

  # Compute variance safely
  vars <- apply(
    df[, feats, drop = FALSE],
    2,
    function(x) var(x, na.rm = TRUE)
  )

  # Keep finite & non-zero variance features
  keep_feats <- names(vars)[is.finite(vars) & vars > 0]

  # ⚠️ If none survive, warn but keep label
  if (length(keep_feats) == 0) {
    warning("All features have zero variance in validation set")
    return(df[, label, drop = FALSE])
  }

  # Final safe subset
  keep_cols <- intersect(c(keep_feats, label), colnames(df))

  df[, keep_cols, drop = FALSE]
}



```


# not working 
```{r}
# --- Function to prepare validation set ---
prepare_validation <- function(physeq_obj, train_mean, train_sd, dc_features) {
  
  # 1️⃣ Extract OTU table and transpose
  otu_df <- as.data.frame(t(otu_table(physeq_obj)))
  
  # 2️⃣ Keep only the features present in the discovery set
  otu_df <- otu_df[, intersect(colnames(otu_df), dc_features), drop = FALSE]
  
  # 3️⃣ Add missing features as zeros
  missing_feats <- setdiff(dc_features, colnames(otu_df))
  if(length(missing_feats) > 0) {
    otu_df[, missing_feats] <- 0
  }
  
  # 4️⃣ Reorder columns to match discovery set
  otu_df <- otu_df[, dc_features]
  
  # 5️⃣ CLR transform
  otu_clr <- microbiome::transform(otu_df, "clr")
  
  # 6️⃣ Z-normalise using discovery mean and sd
  otu_clr_z <- scale(otu_clr, center = train_mean, scale = train_sd)
  
  # 7️⃣ Extract metadata and recode labels
  meta <- data.frame(sample_data(physeq_obj))
  meta$DiseaseStatus <- ifelse(meta$DiseaseStatus %in% c("LOS", "Gut-derived LOS", "1"),
                               "LOS",
                               "Control")
  meta$DiseaseStatus <- factor(meta$DiseaseStatus, levels = c("Control", "LOS"))
  
  # 8️⃣ Combine OTU + labels
  df <- as.data.frame(otu_clr_z) %>%
        mutate(DiseaseStatus = meta$DiseaseStatus)
  
  return(df)
}

# --- Prepare SV1 validation set ---
sv1_df <- prepare_validation(physeqSV1, train_mean, train_sd, colnames(otu_full))

# --- Prepare SV2 validation set ---
sv2_df <- prepare_validation(physeqSV2, train_mean, train_sd, colnames(otu_full))

nrow(sv1_df); nrow(sv2_df)
```
```{r}
dim(sv1_df)           # rows = samples, cols = features+1
apply(sv1_df[, -ncol(sv1_df)], 2, var)  # check for zero variance columns
# keep only features with variance > 0
sv1_df <- sv1_df[, c(apply(sv1_df[, -ncol(sv1_df)], 2, var) > 0, TRUE)]

```

## wrong
```{r}

# Extract CV predictions
pred <- rf_model$pred

# Ensure correct factor levels
pred$obs <- factor(pred$obs, levels = c("Control", "LOS"))

# Predicted probabilities
sv1_prob <- predict(rf_model, sv1_df, type = "prob")[, "LOS"]

# ROC + AUC
roc_sv1 <- roc(
  response  = sv1_df$DiseaseStatus,
  predictor = sv1_prob,
  levels    = c("Control", "LOS"),
  direction = "<"
)

auc_sv1 <- auc(roc_sv1)

# Plot
ggroc(roc_sv1, size = 1.2) +
  theme_minimal() +
  ggtitle(paste0(
    "SV1 Validation Cohort\nAUC = ",
    round(auc_sv1, 3)
  )) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate"
  )

# Predicted probabilities
sv2_prob <- predict(rf_model, sv2_df, type = "prob")[, "LOS"]

# ROC + AUC
roc_sv2 <- roc(
  response  = sv2_df$DiseaseStatus,
  predictor = sv2_prob,
  levels    = c("Control", "LOS"),
  direction = "<"
)

auc_sv2 <- auc(roc_sv2)

# Plot
ggroc(roc_sv2, size = 1.2) +
  theme_minimal() +
  ggtitle(paste0(
    "SV2 Validation Cohort\nAUC = ",
    round(auc_sv2, 3)
  )) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate"
  )


# Predictions on SV1
sv1_pred <- predict(rf_model, sv1_df, type = "prob")
sv1_class <- predict(rf_model, sv1_df)

# Predictions on SV2
sv2_pred <- predict(rf_model, sv2_df, type = "prob")
sv2_class <- predict(rf_model, sv2_df)

library(pROC)  # for ROC/AUC

# --- Function to compute ROC/AUC ---
compute_roc <- function(model, df) {
  # Get predicted probabilities for "LOS"
  probs <- predict(model, df, type = "prob")[, "LOS"]
  
  # True labels (0/1 as factor)
  truth <- df$DiseaseStatus
  
  # ROC object
  roc_obj <- roc(response = truth,
                 predictor = probs,
                 levels = c("Control", "LOS"),  # reference first
                 direction = "<")
  
  # AUC
  auc_val <- auc(roc_obj)
  
  # Plot
  plot(roc_obj, main = paste0("ROC curve (AUC = ", round(auc_val, 3), ")"))
  
  return(list(roc = roc_obj, auc = auc_val))
}

# --- SV1 ROC/AUC ---
sv1_roc <- compute_roc(rf_model, sv1_df)

# --- SV2 ROC/AUC ---
sv2_roc <- compute_roc(rf_model, sv2_df)


```



# try new roc working


```{r}

sv1_plot <- ggroc(
  roc_sv1,
  size = 1.2,
  legacy.axes = TRUE
) +
  theme_minimal() +
  ggtitle(paste0(
    "SV1 Validation Cohort\nAUC = ",
    round(auc_sv1, 3)
  )) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey50"
  ) +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

print(sv1_plot)
ggsave(
  filename = "6b.ROC_Curve_SV1.pdf",
  plot = sv1_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)

###
sv2_plot <- ggroc(
  roc_sv2,
  size = 1.2,
  legacy.axes = TRUE
) +
  theme_minimal() +
  ggtitle(paste0(
    "SV2 Validation Cohort\nAUC = ",
    round(auc_sv2, 3)
  )) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey50"
  ) +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

ggsave(
  filename = "6c.ROC_Curve_SV2.pdf",
  plot = sv2_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)

```


## PR AUC for SV1/2


```{r}

library(PRROC)

# =========================================================
# SV1 PR-AUC
# =========================================================

# Predictions
sv1_prob <- predict(rf_model, sv1_df, type = "prob")[, "LOS"]

sv1_truth <- ifelse(sv1_df$DiseaseStatus == "LOS", 1, 0)

# PR curve
pr_sv1 <- PRROC::pr.curve(
  scores.class0 = sv1_prob[sv1_truth == 1],
  scores.class1 = sv1_prob[sv1_truth == 0],
  curve = TRUE
)

pr_auc_sv1 <- pr_sv1$auc.integral
baseline_sv1 <- mean(sv1_truth)

print(paste("SV1 PR-AUC:", round(pr_auc_sv1, 3)))


# =========================================================
# SV2 PR-AUC
# =========================================================

# Predictions
sv2_prob <- predict(rf_model, sv2_df, type = "prob")[, "LOS"]

sv2_truth <- ifelse(sv2_df$DiseaseStatus == "LOS", 1, 0)

# PR curve
pr_sv2 <- PRROC::pr.curve(
  scores.class0 = sv2_prob[sv2_truth == 1],
  scores.class1 = sv2_prob[sv2_truth == 0],
  curve = TRUE
)

pr_auc_sv2 <- pr_sv2$auc.integral
baseline_sv2 <- mean(sv2_truth)

print(paste("SV2 PR-AUC:", round(pr_auc_sv2, 3)))

pdf("PR_Curve_SV1.pdf", width = 8, height = 6)

plot(
  pr_sv1,
  auc.main = FALSE,
  main = paste0("SV1 PR-AUC = ", round(pr_auc_sv1, 3))
)

abline(h = baseline_sv1, col = "red", lty = 2)

dev.off()

pdf("PR_Curve_SV2.pdf", width = 8, height = 6)

plot(
  pr_sv2,
  auc.main = FALSE,
  main = paste0("SV2 PR-AUC = ", round(pr_auc_sv2, 3))
)

abline(h = baseline_sv2, col = "red", lty = 2)

dev.off()

```



```{r}

# -------------------------
# SV2 predictions
# -------------------------
sv2_df_clean <- sv2_df[complete.cases(sv2_df), ]

sv2_pred <- predict(rf_model, sv2_df_clean, type = "prob")
sv2_prob <- sv2_pred[, "Ecoli_LOS"]

sv2_truth <- ifelse(sv2_df_clean$Ecoli_Status == "Ecoli_LOS", 1, 0)

# -------------------------
# PR curve + PR-AUC
# -------------------------
pr_sv2 <- PRROC::pr.curve(
  scores.class0 = sv2_prob[sv2_truth == 1],
  scores.class1 = sv2_prob[sv2_truth == 0],
  curve = TRUE
)

pr_auc_sv2 <- pr_sv2$auc.integral
print(paste("SV2 PR-AUC:", round(pr_auc_sv2, 3)))

baseline_sv2 <- mean(sv2_truth)

pdf("PR_Curve_SV2.pdf", width = 8, height = 6)

plot(
  pr_sv2,
  auc.main = FALSE,
  main = paste0(
    "SV2 PR Curve\nPR-AUC = ",
    round(pr_auc_sv2, 3),
    " | baseline = ",
    round(baseline_sv2, 3)
  )
)

abline(h = baseline_sv2, col = "red", lty = 2)

dev.off()

```



```{r}
library(caret)

# --- SV1 Metrics ---
sv1_cm <- confusionMatrix(sv1_class, sv1_df$DiseaseStatus, positive = "LOS")
sv1_cm
```
```{r}
# --- SV2 Metrics ---
sv2_cm <- confusionMatrix(sv2_class, sv2_df$DiseaseStatus, positive = "LOS")# 
sv2_cm
```


```{r}
library(caret)
library(pROC)
library(MLmetrics)

# Predicted classes and probabilities
sv1_class <- predict(rf_model, sv1_df)
sv1_prob  <- predict(rf_model, sv1_df, type = "prob")[, "LOS"]  # probability for LOS

# Confusion Matrix
cm <- confusionMatrix(sv1_class, sv1_df$DiseaseStatus, positive = "LOS")

# Extract Accuracy, Sensitivity, Specificity
accuracy    <- cm$overall["Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]

# Precision
precision <- Precision(y_pred = sv1_class, y_true = sv1_df$DiseaseStatus, positive = "LOS")

# F1 Score
f1 <- F1_Score(y_pred = sv1_class, y_true = sv1_df$DiseaseStatus, positive = "LOS")

# AUC
roc_obj <- roc(response = sv1_df$DiseaseStatus,
               predictor = sv1_prob,
               levels = c("Control","LOS"), direction = "<")
auc_val <- auc(roc_obj)

# Print all
metrics <- data.frame(
  Accuracy    = accuracy,
  Sensitivity = sensitivity,
  Specificity = specificity,
  Precision   = precision,
  F1          = f1,
  AUC         = as.numeric(auc_val)
)
print(metrics)

```


```{r}
library(caret)
library(pROC)
library(MLmetrics)

# Predicted classes and probabilities
sv2_class <- predict(rf_model, sv2_df)
sv2_prob  <- predict(rf_model, sv2_df, type = "prob")[, "LOS"]  # probability for LOS

# Confusion Matrix
cm <- confusionMatrix(sv2_class, sv2_df$DiseaseStatus, positive = "LOS")

# Extract Accuracy, Sensitivity, Specificity
accuracy    <- cm$overall["Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]

# Precision
precision <- Precision(y_pred = sv2_class, y_true = sv2_df$DiseaseStatus, positive = "LOS")

# F1 Score
f1 <- F1_Score(y_pred = sv2_class, y_true = sv2_df$DiseaseStatus, positive = "LOS")

# AUC
roc_obj <- roc(response = sv2_df$DiseaseStatus,
               predictor = sv2_prob,
               levels = c("Control","LOS"), direction = "<")
auc_val <- auc(roc_obj)

# Print all
metrics <- data.frame(
  Accuracy    = accuracy,
  Sensitivity = sensitivity,
  Specificity = specificity,
  Precision   = precision,
  F1          = f1,
  AUC         = as.numeric(auc_val)
)
print(metrics)

```
```{r}
common_species

```

