
# 

#Load libraries and data:

Required libraries:
```{r}
library(vegan)
library (ggplot2)
library(data.table)
library(cowplot)
#library(Maaslin2)
library(readr)
library(phyloseq)
library(pgirmess)
library(agricolae)
library(readxl) 
library(caret)
library(tidyverse)
library(microbiome)
library(MLmetrics)
library(randomForest)
library(gridExtra)
library(grid)
library(pROC)  # for ROC/AUC

```

#How to make phyloseq object:

Reformat table:


```{r}

otufile = "20250512_OTU tables_percohort_sepsis prediction model.xlsx"
metafile = "20260326_Overview samples for Animesh_addedclinicalparameters.xlsx"

# ------------------------------------------------------------
# 2️⃣ Read OTU table (Discovery cohort - DC)
# ------------------------------------------------------------
discDC <- read_excel(otufile,
                     sheet = "DC") %>% as.data.frame()

# First column contains sample IDs → make them rownames
rownames(discDC) <- discDC[[1]]
discDC <- discDC[ , -1]

# Remove unwanted PC samples (rows)
remove_samples <- paste0("DC-PC", 1:13)
discDC <- discDC[!rownames(discDC) %in% remove_samples, ]
cat("Number of samples after removing PC samples:", nrow(discDC), "\n")

# Transpose: rows = OTUs, cols = samples
DCotumat <- t(discDC)

# ------------------------------------------------------------
# 3️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel(otufile,
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# Keep only OTUs present in both OTU and taxonomy tables
common_otus <- intersect(rownames(DCotumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# Subset OTU and taxonomy tables
DCotumat_common <- DCotumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# Sanity check
dim(DCotumat_common)
dim(taxonomy_common)

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(DCotumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqDC <- phyloseq(OTU, TAX)
cat("physeqDC - Samples:", nsamples(physeqDC), "OTUs:", ntaxa(physeqDC), "\n")

## new added
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel(metafile, 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Discovery cohort", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"

# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(DCotumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqDC <- merge_phyloseq(physeqDC, sampledata)

cat("physeqDC with metadata - Samples:", nsamples(physeqDC), 
    "OTUs:", ntaxa(physeqDC), "\n")


```


# If you also have SV1 and SV2 phyloseq objects

# prep SV1
```{r}

# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV1
# ------------------------------------------------------------
discSV1 <- read_excel(otufile, 
                      sheet = "SV 1") %>% as.data.frame()

# first column = sample IDs
rownames(discSV1) <- discSV1[[1]]
discSV1 <- discSV1[ , -1]

# transpose: rows = OTUs, cols = samples
SV1otumat <- t(discSV1)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel(otufile,
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV1otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# subset OTU and taxonomy tables
SV1otumat_common <- SV1otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV1otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV1 <- phyloseq(OTU, TAX)

cat("physeqSV1 - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel(metafile, 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 1", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"
# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV1otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV1 <- merge_phyloseq(physeqSV1, sampledata)

cat("physeqSV1 with metadata - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

```

# prep SV2 data
```{r}
# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV2
# ------------------------------------------------------------
discSV2 <- read_excel(otufile, 
                      sheet = "SV 2") %>% as.data.frame()

# Suppose the first column contains the row IDs (sample names)
rownames(discSV2) <- discSV2[[1]]
discSV2 <- discSV2[ , -1]  # remove the first column after assigning rownames

# Now remove rows that contain "SV2-Blanco"
discSV2 <- discSV2[!grepl("^SV2-Blanco", rownames(discSV2)), ]

cat("Number of rows after removing blanks:", nrow(discSV2), "\n")

# transpose: rows = OTUs, cols = samples
SV2otumat <- t(discSV2)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel(otufile,
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV2otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

SV2otumat_common <- SV2otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV2otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV2 <- phyloseq(OTU, TAX)

cat("physeqSV2 - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel(metafile, 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 2
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 2", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"
# Keep only rows that exist in SV2 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV2otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV2 <- merge_phyloseq(physeqSV2, sampledata)

cat("physeqSV2 with metadata - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

```

# til sv2 prep 

## newruns for Model for E.coli sepsis with overlapping ASVX 

```{r}
# ------------------------------------------------------------
# Compute common OTUs across Discovery (DC), VC1, VC2 cohorts
# ------------------------------------------------------------

# Extract OTU tables (no species collapsing)
otu_full <- as.data.frame(t(otu_table(physeqDC)))    # Discovery cohort
otu_sv1  <- as.data.frame(t(otu_table(physeqSV1)))   # Validation 1
otu_sv2  <- as.data.frame(t(otu_table(physeqSV2)))   # Validation 2

# Find OTUs present in all three cohorts
common_otus <- Reduce(intersect, list(
  colnames(otu_full),
  colnames(otu_sv1),
  colnames(otu_sv2)
))

cat("Number of common OTUs:", length(common_otus), "\n")


```

# E. coli LOS vs Control (Discovery Cohort, all time periods)


```{r}

# ------------------------------------------------------------
# 2️⃣ Prepare metadata for E. coli LOS vs Control comparison
# ------------------------------------------------------------
meta_ecoli <- data.frame(sample_data(physeqDC))

# Recode DiseaseStatus numeric to text
meta_ecoli$DiseaseStatus_txt <- ifelse(meta_ecoli$DiseaseStatus == 1, "LOS", "Control")

# Define Ecoli_Status
meta_ecoli$Ecoli_Status <- NA

# Label E. coli LOS cases
meta_ecoli$Ecoli_Status[
  meta_ecoli$Bloodculture_isolated_pathogen == "Escherichia coli" &
  meta_ecoli$DiseaseStatus_txt == "LOS"
] <- "Ecoli_LOS"

# Label Controls (all controls, regardless of pathogen)
meta_ecoli$Ecoli_Status[
  meta_ecoli$DiseaseStatus_txt == "Control"
] <- "Control"

# Check distribution
table(meta_ecoli$Ecoli_Status)


# Subset phyloseq to only these samples
physeq_ecoli <- prune_samples(rownames(meta_ecoli), physeqDC)


# ------------------------------------------------------------
# 3️⃣ Extract OTU table (common OTUs, not species)
# ------------------------------------------------------------
otu_ecoli <- as.data.frame(t(otu_table(physeq_ecoli)))

# Filter to only common OTUs used across cohorts
otu_ecoli <- otu_ecoli[, common_otus, drop = FALSE]
  # reuse from previous step (common_species == common OTUs)

# Add pseudocount and apply CLR + Z-normalization
otu_ecoli_pseudo <- otu_ecoli + 1
otu_ecoli_clr <- microbiome::transform(otu_ecoli_pseudo, "clr")
otu_ecoli_clr_z <- scale(otu_ecoli_clr)

# Save scaling parameters for validation
train_mean <- attr(otu_ecoli_clr_z, "scaled:center")
train_sd   <- attr(otu_ecoli_clr_z, "scaled:scale")


# ------------------------------------------------------------
# 4️⃣ Combine with metadata
# ------------------------------------------------------------
df_ecoli <- as.data.frame(otu_ecoli_clr_z)
df_ecoli$Ecoli_Status <- factor(meta_ecoli$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

cat("Number of predictors:", ncol(df_ecoli) - 1, "\n")
cat("Samples per class:\n")
print(table(df_ecoli$Ecoli_Status))

```

```{r}

# ------------------------------------------------------------
# 1️⃣ Ensure no missing values
# ------------------------------------------------------------
# Remove samples with NA in Ecoli_Status
# Keep only samples with known Ecoli_Status
df_ecoli_clean <- df_ecoli[!is.na(df_ecoli$Ecoli_Status), ]

# Ensure factor levels are correct
df_ecoli_clean$Ecoli_Status <- factor(df_ecoli_clean$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

# Optional: replace any remaining NA in OTUs
df_ecoli_clean[is.na(df_ecoli_clean)] <- 0

# Check distribution
table(df_ecoli_clean$Ecoli_Status)

# 2️⃣ Split into training/testing
set.seed(123)
train_index <- createDataPartition(df_ecoli_clean$Ecoli_Status, p = 0.8, list = FALSE)
train_set   <- df_ecoli_clean[train_index, ]
test_set    <- df_ecoli_clean[-train_index, ]

# ------------------------------------------------------------
# 3️⃣ Custom summary function for caret
# ------------------------------------------------------------
multiSummary <- function(data, lev = NULL, model = NULL) {
  roc_val <- tryCatch({
    roc(response = data$obs, predictor = data[, lev[2]], levels = rev(lev), quiet = TRUE)$auc
  }, error = function(e) NA)
  
  out <- c(
    ROC       = roc_val,
    Sens      = sensitivity(data$pred, data$obs, positive = lev[2]),
    Spec      = specificity(data$pred, data$obs, negative = lev[1]),
    Precision = Precision(y_pred = data$pred, y_true = data$obs, positive = lev[2]),
    Recall    = Recall(y_pred = data$pred, y_true = data$obs, positive = lev[2]),
    F1        = F1_Score(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  )
  return(out)
}


# ------------------------------------------------------------
# Train Random Forest (ranger)
# ------------------------------------------------------------
set.seed(123)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 3,     # smaller folds for rare class else use 10
  repeats = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  sampling = "up",   # upsample minority class
  savePredictions = "final"
)

#  Check class distribution
table(train_set$Ecoli_Status)
table(test_set$Ecoli_Status)

rf_model <- train(
  Ecoli_Status ~ .,
  data       = df_ecoli_clean,
  method     = "ranger",
  metric     = "F1",          # or "Accuracy"
  tuneGrid   = expand.grid(
                 mtry          = floor(sqrt(ncol(df_ecoli_clean) - 1)),
                 splitrule     = "gini",
                 min.node.size = 1
               ),
  trControl  = ctrl,
  importance = "impurity"
)



library(caret)
library(MLmetrics)
library(pROC)
library(dplyr)

# ------------------------------------------------------------
# Extract predictions from the trained model
# ------------------------------------------------------------
pred <- rf_model$pred

# Ensure factor levels
pred$obs <- factor(pred$obs, levels = c("Control", "Ecoli_LOS"))
pred$pred <- factor(pred$pred, levels = c("Control", "Ecoli_LOS"))

# ------------------------------------------------------------
# Compute metrics per fold
# ------------------------------------------------------------
fold_metrics <- pred %>%
  group_by(Resample) %>%
  summarise(
    Precision   = Precision(y_pred = pred, y_true = obs, positive = "Ecoli_LOS"),
    F1_score    = F1_Score(y_pred = pred, y_true = obs, positive = "Ecoli_LOS"),
    Sensitivity = Sensitivity(y_pred = pred, y_true = obs, positive = "Ecoli_LOS"),
    Specificity = Specificity(y_pred = pred, y_true = obs, positive = "Ecoli_LOS"),
    AUROC       = as.numeric(auc(roc(response = obs, predictor = Ecoli_LOS, levels = c("Control","Ecoli_LOS"))))
  )

# ------------------------------------------------------------
# Compute AVERAGE and SD across folds
# ------------------------------------------------------------
avg_row <- colMeans(fold_metrics[,2:6], na.rm = TRUE)
sd_row  <- apply(fold_metrics[,2:6], 2, sd, na.rm = TRUE)

summary_table <- rbind(AVERAGE = avg_row, SD = sd_row)
summary_table <- as.data.frame(summary_table)

print(summary_table)



```

# extra for ROC
```{r}
library(pROC)
library(ggplot2)
library(dplyr)

# Ensure correct factor levels
pred <- rf_model$pred
pred$obs <- factor(pred$obs, levels = c("Control", "Ecoli_LOS"))

# Compute ROC
roc_discovery <- roc(
  response  = pred$obs,
  predictor = pred$Ecoli_LOS,
  levels    = c("Control", "Ecoli_LOS"),
  direction = "<"
)

auc_val <- auc(roc_discovery)

# Create the plot
roc_plot <- ggroc(roc_discovery, legacy.axes = TRUE,
                  colour = "black", size = 1.2) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dotted", colour = "grey50") +
  ggtitle(paste0("ROC Curve – Discovery Cohort (CV)\nAUC = ",
                 round(auc_val, 3))) +
  theme_minimal() +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

# Show the plot in R
print(roc_plot)

# Save as PDF
ggsave(
  filename = "MLfiles/6d.ROC_Curve_Discovery.pdf",
  plot = roc_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)


```


## PR-AUC for Discovery Cohort (CV)

```{r}

library(PRROC)

pred <- rf_model$pred

# Ensure correct ordering
pred$obs <- factor(pred$obs, levels = c("Control", "Ecoli_LOS"))

# True labels (1 = positive)
truth_dc <- ifelse(pred$obs == "Ecoli_LOS", 1, 0)

# Predicted probabilities
prob_dc <- pred$Ecoli_LOS

pr_dc <- pr.curve(
  scores.class0 = prob_dc[truth_dc == 1],  # positives
  scores.class1 = prob_dc[truth_dc == 0],  # negatives
  curve = TRUE
)

pr_auc_dc <- pr_dc$auc.integral
print(paste("Discovery PR-AUC:", round(pr_auc_dc, 3)))

baseline_dc <- mean(truth_dc)

plot(
  pr_dc,
  auc.main = FALSE,
  main = paste("Discovery PR Curve (CV)\nAUC =",
               round(pr_auc_dc, 2),
               ", baseline =",
               round(baseline_dc, 2))
)

abline(h = baseline_dc, col = "red", lty = 2)


pdf("MLfiles/E. coli-LOS model PR_Curve_Discovery.pdf", width = 8, height = 6)

plot(
  pr_dc,
  auc.main = FALSE,
  main = paste("Discovery PR Curve (CV)\nAUC =",
               round(pr_auc_dc, 2),
               ", baseline =",
               round(baseline_dc, 2))
)

abline(h = baseline_dc, col = "red", lty = 2)

dev.off()



```



## Out-of-Bag (OOB) error
```{r}

rf_model$finalModel$prediction.error

```


###  end this DC ML ###





# only for DC - top 20 otus

```{r}

# Extract variable importance
var_imp <- varImp(rf_model)$importance

# Add Original_ID column directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending
var_imp_ordered <- var_imp[order(-var_imp$Overall), ]

# Select top 20
top20 <- head(var_imp_ordered, 20)

# Plot
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()


```


# change every time file name 
```{r}

# --- 1️⃣ Extract variable importance ---
var_imp <- varImp(rf_model)$importance

# Add Original_ID directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending and add Rank
var_imp <- var_imp[order(-var_imp$Overall), ]
var_imp$Rank <- 1:nrow(var_imp)

# Top 20
top20 <- head(var_imp, 20)

# --- 2️⃣ Plot top 20 OTUs ---
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()

# --- 3️⃣ Merge taxonomy info ---
tax <- read_excel(
  otufile,
  sheet = "OTU table all"
) %>%
  mutate(ID.1 = as.character(ID.1))

top20 <- top20 %>%
  mutate(Original_ID = as.character(Original_ID)) %>%
  left_join(tax %>% select(ID.1, Family, Genus, Species),
            by = c("Original_ID" = "ID.1")) %>%
  select(Rank, Original_ID, Family, Genus, Species, Overall)

# --- 4️⃣ Create table grob ---
tab_grob <- tableGrob(
  top20,
  rows = NULL,
  theme = ttheme_default(
    core   = list(fg_params = list(cex = 0.8)),
    colhead= list(fg_params = list(cex = 1, fontface = "bold"))
  )
)

title <- textGrob("Top 20 OTUs with Taxonomy",
                  gp = gpar(fontsize = 16, fontface = "bold"))

# --- 5️⃣ Save as PNG ---
png("figs/ecoli_Top20_OTU_Taxonomy.png", width = 2000, height = 1400, res = 200)
grid.arrange(title, tab_grob, ncol = 1, heights = c(0.1, 0.8, 0.1))
dev.off()

```

## new mod for f_Mitochondria 

```{r}

# --- 1️⃣ Extract variable importance ---
var_imp <- varImp(rf_model)$importance
var_imp$Original_ID <- rownames(var_imp)

# Order by importance
var_imp <- var_imp[order(-var_imp$Overall), ]

# --- 2️⃣ Merge taxonomy FIRST ---
tax <- read_excel(
  otufile,
  sheet = "OTU table all"
) %>%
  mutate(ID.1 = as.character(ID.1))

var_imp <- var_imp %>%
  mutate(Original_ID = as.character(Original_ID)) %>%
  left_join(tax %>% select(ID.1, Family, Genus, Species),
            by = c("Original_ID" = "ID.1"))

# --- 3️⃣ Remove NA taxonomy (KEY STEP) ---
var_imp <- var_imp %>%
  filter(!is.na(Genus))

# --- 4️⃣ Re-rank AFTER filtering ---
var_imp$Rank <- 1:nrow(var_imp)

# --- 5️⃣ Take top 20 valid taxa ---
ntop20 <- head(var_imp, 20)


library(gridExtra)
library(grid)

# --- 4️⃣ Create table grob ---
tab_grob <- tableGrob(
  ntop20,
  rows = NULL,
  theme = ttheme_default(
    core   = list(fg_params = list(cex = 0.8)),
    colhead= list(fg_params = list(cex = 1, fontface = "bold"))
  )
)

# --- Title ---
title <- textGrob(
  "Top 20 OTUs with Taxonomy",
  gp = gpar(fontsize = 16, fontface = "bold")
)

# --- 5️⃣ Save as PNG ---
png("figs/new-ecoli_Top20_OTU_Taxonomy.png",
    width = 2000,
    height = 1400,
    res = 200)

grid.arrange(
  title,
  tab_grob,
  ncol = 1,
  heights = c(0.1, 0.9)   # ✅ FIX: remove extra unused space
)

dev.off()

##ggsave("figs/new-ecoli_Top20_OTU_Taxonomy.png", plot = p, width = 8, height = 6, dpi = 300)

```




# now test in SVs

# --- Function to prepare validation set ---
```{r}
# --- Function to prepare validation set ---

prepare_validation <- function(physeq_obj, train_mean, train_sd, dc_features) {
  
  # 1️⃣ Extract OTU table and transpose
  otu_df <- as.data.frame(t(otu_table(physeq_obj)))
  
  # 2️⃣ Remove duplicate OTU names (if any)
  otu_df <- otu_df[, !duplicated(colnames(otu_df)), drop = FALSE]
  
  # 3️⃣ Add missing OTUs as zeros
  missing_feats <- setdiff(dc_features, colnames(otu_df))
  if(length(missing_feats) > 0) {
    warning("Adding missing OTUs as zeros: ", paste(missing_feats, collapse = ", "))
    otu_df[, missing_feats] <- 0
  }
  
  # 4️⃣ Keep only DC OTUs and reorder exactly
  otu_df <- otu_df[, dc_features, drop = FALSE]
  
  # 5️⃣ Add pseudocount
  otu_df <- otu_df + 1
  
  # 6️⃣ CLR transform
  log_mat <- log(otu_df)
  row_means <- rowMeans(log_mat)
  otu_clr <- sweep(log_mat, 1, row_means, "-")
  
  # 7️⃣ Check dimensions before scaling
  if(ncol(otu_clr) != length(train_mean) || ncol(otu_clr) != length(train_sd)) {
    stop(
      "Mismatch in number of columns: ",
      "OTU table has ", ncol(otu_clr), 
      ", train_mean has ", length(train_mean),
      ", train_sd has ", length(train_sd)
    )
  }
  
  # 8️⃣ Z-normalize
  otu_clr_z <- scale(otu_clr, center = train_mean, scale = train_sd)
  
  # 9️⃣ Metadata & labels
  meta <- data.frame(sample_data(physeq_obj))
  meta$Ecoli_Status <- NA
  meta$Ecoli_Status[
    meta$Bloodculture_isolated_pathogen == "Escherichia coli" &
    meta$DiseaseStatus %in% c(1, "LOS")
  ] <- "Ecoli_LOS"
  meta$Ecoli_Status[
    meta$DiseaseStatus %in% c(0, "Control")
  ] <- "Control"
  
  meta$Ecoli_Status <- factor(meta$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))
  
  # 🔟 Combine OTU + labels
  df <- as.data.frame(otu_clr_z)
  df$Ecoli_Status <- meta$Ecoli_Status
  df <- df[!is.na(df$Ecoli_Status), ]
  
  return(df)
}


# Use only OTU columns, exclude Ecoli_Status
dc_features <- setdiff(colnames(df_ecoli), "Ecoli_Status")

sv1_df <- prepare_validation(physeqSV1, train_mean, train_sd, dc_features)
sv2_df <- prepare_validation(physeqSV2, train_mean, train_sd, dc_features)

nrow(sv1_df); nrow(sv2_df)
```

# try new roc

```{r}

library(pROC)

roc_dc <- roc(
  response  = pred$obs,
  predictor = pred$Ecoli_LOS,
  levels    = c("Control", "Ecoli_LOS"),
  direction = "<"
)
auc_dc <- auc(roc_dc)

opt_thresh_dc <- coords(
  roc_dc,
  x = "best",
  best.method = "youden",
  ret = "threshold"
)

print(opt_thresh_dc)

sv1_prob <- predict(rf_model, sv1_df, type = "prob")[, "Ecoli_LOS"]

roc_sv1 <- roc(
  response  = sv1_df$Ecoli_Status,
  predictor = sv1_prob,
  levels    = c("Control", "Ecoli_LOS"),
  direction = "<"
)
auc_sv1 <- auc(roc_sv1)

sv2_prob <- predict(rf_model, sv2_df, type = "prob")[, "Ecoli_LOS"]

roc_sv2 <- roc(
  response  = sv2_df$Ecoli_Status,
  predictor = sv2_prob,
  levels    = c("Control", "Ecoli_LOS"),
  direction = "<"
)
auc_sv2 <- auc(roc_sv2)


library(ggplot2)

ggroc(roc_sv1, size = 1.2) +
  theme_minimal() +
  ggtitle(paste0("roc_sv1 (CV)\nAUC = ", round(auc_sv1, 3))) +
  labs(x = "False Positive Rate", y = "True Positive Rate")

```


## 
```{r}
#### precision–recall AUC
library(PRROC)
library(pROC)
library(MLmetrics)
library(caret)

# ------------------------------------------------------------
# 1️⃣ Clean SV1 data
# ------------------------------------------------------------
sv1_df_clean <- sv1_df[complete.cases(sv1_df), ]

# ------------------------------------------------------------
# 2️⃣ Predictions
# ------------------------------------------------------------
sv1_pred <- predict(rf_model, sv1_df_clean, type = "prob")
sv1_prob <- sv1_pred[, "Ecoli_LOS"]

# ------------------------------------------------------------
# 3️⃣ True labels
# ------------------------------------------------------------
sv1_truth <- ifelse(sv1_df_clean$Ecoli_Status == "Ecoli_LOS", 1, 0)

# ------------------------------------------------------------
# 4️⃣ PR curve
# ------------------------------------------------------------
pr_sv1 <- pr.curve(
  scores.class0 = sv1_prob[sv1_truth == 1],
  scores.class1 = sv1_prob[sv1_truth == 0],
  curve = TRUE
)

# ------------------------------------------------------------
# 5️⃣ PR-AUC
# ------------------------------------------------------------
pr_auc_sv1 <- pr_sv1$auc.integral
print(paste("SV1 PR-AUC:", round(pr_auc_sv1, 3)))

# ------------------------------------------------------------
# 6️⃣ Baseline
# ------------------------------------------------------------
baseline_sv1 <- mean(sv1_truth)

# ------------------------------------------------------------
# 7️⃣ Get threshold from DC (IMPORTANT)
# ------------------------------------------------------------
opt_thresh_dc <- coords(
  roc_dc,
  x = "best",
  best.method = "youden",
  ret = "threshold"
)

print(paste("Optimal threshold (DC):", round(opt_thresh_dc, 3)))

# ------------------------------------------------------------
# 8️⃣ Classification using DC threshold
# ------------------------------------------------------------
sv1_class <- factor(
  ifelse(sv1_prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# ------------------------------------------------------------
# 9️⃣ F1-score
# ------------------------------------------------------------
# ------------------------------------------------------------
# Create evaluation dataframe (SAFE)
# ------------------------------------------------------------
df_eval_sv1 <- data.frame(
  prob  = sv1_prob,
  truth = sv1_df_clean$Ecoli_Status
)

# Remove any remaining NA
df_eval_sv1 <- df_eval_sv1[complete.cases(df_eval_sv1), ]

# ------------------------------------------------------------
# Classification using DC threshold
# ------------------------------------------------------------
df_eval_sv1$class <- factor(
  ifelse(df_eval_sv1$prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# ------------------------------------------------------------
# F1-score (NOW SAFE)
# ------------------------------------------------------------
library(MLmetrics)

F1_sv1 <- F1_Score(
  y_pred = df_eval_sv1$class,
  y_true = factor(df_eval_sv1$truth),
  positive = "Ecoli_LOS"
)

if (is.nan(F1_sv1)) F1_sv1 <- 0

print(paste("SV1 F1-score:", round(F1_sv1, 3)))

# ------------------------------------------------------------
# 🔟 Show plot in R
# ------------------------------------------------------------
plot(
  pr_sv1,
  auc.main = FALSE,
  main = paste("SV1 PR Curve (AUC =", round(pr_auc_sv1, 2),
               ", baseline =", round(baseline_sv1, 2), ")")
)

abline(h = baseline_sv1, col = "red", lty = 2)

# ------------------------------------------------------------
# 1️⃣1️⃣ Save PNG
# ------------------------------------------------------------
png("figs/SV1_PR_curve.png", width = 2000, height = 1200, res = 300)

plot(
  pr_sv1,
  auc.main = FALSE,
  main = paste("SV1 PR Curve (AUC =", round(pr_auc_sv1, 2),
               ", baseline =", round(baseline_sv1, 2),
                ")")
)

abline(h = baseline_sv1, col = "red", lty = 2)

dev.off()
```


```{r}

library(PRROC)

# ------------------------------------------------------------
# 1️⃣ Clean SV2 data
# ------------------------------------------------------------
sv2_df_clean <- sv2_df[complete.cases(sv2_df), ]

# ------------------------------------------------------------
# 2️⃣ Predictions
# ------------------------------------------------------------
sv2_pred <- predict(rf_model, sv2_df_clean, type = "prob")
sv2_prob <- sv2_pred[, "Ecoli_LOS"]

# ------------------------------------------------------------
# 3️⃣ True labels
# ------------------------------------------------------------
sv2_truth <- ifelse(sv2_df_clean$Ecoli_Status == "Ecoli_LOS", 1, 0)

# ------------------------------------------------------------
# 4️⃣ PR curve
# ------------------------------------------------------------
pr_sv2 <- pr.curve(
  scores.class0 = sv2_prob[sv2_truth == 1],   # positives
  scores.class1 = sv2_prob[sv2_truth == 0],   # negatives
  curve = TRUE
)

# ------------------------------------------------------------
# 5️⃣ PR-AUC
# ------------------------------------------------------------
pr_auc_sv2 <- pr_sv2$auc.integral
print(pr_auc_sv2)

# ------------------------------------------------------------
# 6️⃣ Baseline (VERY IMPORTANT)
# ------------------------------------------------------------
baseline_sv2 <- mean(sv2_truth)

# File name
png("figs/SV2_PR_curve.png", width = 2000, height = 1200, res = 300)

# ------------------------------------------------------------
# 7️⃣ Plot (clean, publication-ready)
# ------------------------------------------------------------
plot(
  pr_sv2,
  auc.main = FALSE,
  main = paste("SV2 PR Curve (AUC =", round(pr_auc_sv2, 2),
               ", baseline =", round(baseline_sv2, 2), ")")
)

abline(h = baseline_sv2, col = "red", lty = 2)
dev.off()
```



```{r}
library(PRROC)
library(caret)

# ------------------------------------------------------------
# 1️⃣ Clean SV1 data
# ------------------------------------------------------------
sv1_df_clean <- sv1_df[complete.cases(sv1_df), ]

# ------------------------------------------------------------
# 2️⃣ Predictions
# ------------------------------------------------------------
sv1_prob <- predict(rf_model, sv1_df_clean, type = "prob")[, "Ecoli_LOS"]

# ------------------------------------------------------------
# 3️⃣ True labels
# ------------------------------------------------------------
sv1_truth <- ifelse(sv1_df_clean$Ecoli_Status == "Ecoli_LOS", 1, 0)

# ------------------------------------------------------------
# 4️⃣ PR curve
# ------------------------------------------------------------
pr_sv1 <- pr.curve(
  scores.class0 = sv1_prob[sv1_truth == 1],
  scores.class1 = sv1_prob[sv1_truth == 0],
  curve = TRUE
)

# ------------------------------------------------------------
# 5️⃣ PR-AUC
# ------------------------------------------------------------
pr_auc_sv1 <- pr_sv1$auc.integral
print(paste("SV1 PR-AUC:", round(pr_auc_sv1, 3)))

# ------------------------------------------------------------
# 6️⃣ Baseline
# ------------------------------------------------------------
baseline_sv1 <- mean(sv1_truth)

# ------------------------------------------------------------
# 7️⃣ Classification (using DC threshold)
# ------------------------------------------------------------
sv1_class <- factor(
  ifelse(sv1_prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# ------------------------------------------------------------
# 8️⃣ Confusion matrix
# ------------------------------------------------------------
# ------------------------------------------------------------
# Create evaluation dataframe (SAFE)
# ------------------------------------------------------------
df_eval_sv1 <- data.frame(
  prob  = sv1_prob,
  truth = sv1_df_clean$Ecoli_Status
)

# Remove any NA rows (important)
df_eval_sv1 <- df_eval_sv1[complete.cases(df_eval_sv1), ]

# ------------------------------------------------------------
# Classification using DC threshold
# ------------------------------------------------------------
df_eval_sv1$class <- factor(
  ifelse(df_eval_sv1$prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# ------------------------------------------------------------
# Confusion matrix (NOW SAFE)
# ------------------------------------------------------------
cm <- confusionMatrix(
  df_eval_sv1$class,
  factor(df_eval_sv1$truth, levels = c("Control", "Ecoli_LOS"))
)

print(cm)

# ------------------------------------------------------------
# 9️⃣ F1-score
# ------------------------------------------------------------
precision <- cm$byClass["Precision"]
recall    <- cm$byClass["Recall"]

F1_sv1 <- 2 * (precision * recall) / (precision + recall)

# Handle NaN safely
if (is.nan(F1_sv1)) F1_sv1 <- 0

print(paste("SV1 F1-score:", round(F1_sv1, 3)))

# ------------------------------------------------------------
# 🔟 Plot (clean + publication-ready)
# ------------------------------------------------------------
plot(
  pr_sv1,
  auc.main = FALSE,
  main = paste("SV1 PR Curve (AUC =", round(pr_auc_sv1, 2),
               ", baseline =", round(baseline_sv1, 2),
               ", F1 =", round(F1_sv1, 2), ")")
)

abline(h = baseline_sv1, col = "red", lty = 2)
```



```{r}

#### precision–recall AUC

library(PRROC)

# SV1 PR-AUC
sv2_prob <- predict(rf_model, sv2_df, type = "prob")[, "Ecoli_LOS"]

# True labels (convert to binary: 1 = positive)
sv2_truth <- ifelse(sv2_df$Ecoli_Status == "Ecoli_LOS", 1, 0)

pr_sv2 <- pr.curve(
  scores.class0 = sv2_prob[sv2_truth == 1],   # positives
  scores.class1 = sv2_prob[sv2_truth == 0],   # negatives
  curve = TRUE
)

# PR-AUC value
pr_sv2$auc.integral

library(caret)

sv1_class <- factor(
  ifelse(sv2_prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# Confusion matrix
# ------------------------------------------------------------
# Create evaluation dataframe (SAFE)
# ------------------------------------------------------------
df_eval_sv2 <- data.frame(
  prob  = sv2_prob,
  truth = sv2_df_clean$Ecoli_Status
)

# Remove any NA rows (important)
df_eval_sv2 <- df_eval_sv2[complete.cases(df_eval_sv2), ]

# ------------------------------------------------------------
# Classification using DC threshold
# ------------------------------------------------------------
df_eval_sv2$class <- factor(
  ifelse(df_eval_sv2$prob >= opt_thresh_dc, "Ecoli_LOS", "Control"),
  levels = c("Control", "Ecoli_LOS")
)

# ------------------------------------------------------------
# Confusion matrix (NOW SAFE)
# ------------------------------------------------------------
cm_sv2 <- confusionMatrix(
  df_eval_sv2$class,
  factor(df_eval_sv2$truth, levels = c("Control", "Ecoli_LOS"))
)

print(cm_sv2)

precision <- cm_sv2$byClass["Precision"]
recall    <- cm_sv2$byClass["Recall"]

F1_sv2 <- 2 * (precision * recall) / (precision + recall)

# Handle NaN safely
if (is.nan(F1_sv2)) F1_sv2 <- 0

print(paste("SV2 F1-score:", round(F1_sv2, 3)))


plot(pr_sv2, main = paste("SV2 PR Curve (AUC =", round(pr_sv2$auc.integral, 3), ")"))

# 9️⃣ Plot with F1
# ------------------------------------------------------------
plot(
  pr_sv2,
  auc.main = FALSE,
  main = paste("SV2 PR Curve (AUC =", round(pr_auc_sv2, 2),
               ", baseline =", round(baseline, 2),
               ", F1 =", round(F1, 2), ")")
)

abline(h = baseline, col = "red", lty = 2)

```


```{r}

F1_Score(
  y_pred = sv1_class,
  y_true = factor(sv1_df_clean$Ecoli_Status),
  positive = "Ecoli_LOS"
)

```


```{r}
sv1_plot <- ggroc(roc_sv1, legacy.axes = TRUE, size = 1.2) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dotted", colour = "grey50") +
  theme_minimal() +
  ggtitle(paste0("SV1 Validation Cohort\nAUC = ", round(auc_sv1, 3))) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

print(sv1_plot)

ggsave(
  filename = "MLfiles/6e.ROC_Curve_SV1.pdf",
  plot = sv1_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)


sv2_plot <- ggroc(roc_sv2, legacy.axes = TRUE, size = 1.2) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dotted", colour = "grey50") +
  theme_minimal() +
  ggtitle(paste0("SV2 Validation Cohort\nAUC = ", round(auc_sv2, 3))) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  coord_equal() +
  xlim(0, 1) +
  ylim(0, 1)

print(sv2_plot)

ggsave(
  filename = "MLfiles/6f.ROC_Curve_SV2.pdf",
  plot = sv2_plot,
  width = 8,
  height = 6,
  device = cairo_pdf
)

```





```{r}
## Make sure the factor levels match
sv1_df$Ecoli_Status <- factor(sv1_df$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))
sv2_df$Ecoli_Status <- factor(sv2_df$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

# Predictions
sv1_class <- predict(rf_model, sv1_df)
sv2_class <- predict(rf_model, sv2_df)

# Confusion matrices
library(caret)
sv1_cm <- confusionMatrix(sv1_class, sv1_df$Ecoli_Status, positive = "Ecoli_LOS")#
sv1_cm


```


```{r}
library(MLmetrics)

# Make sure levels match
sv1_df$Ecoli_Status <- factor(sv1_df$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

# Precision
Precision_val <- Precision(y_pred = sv1_class, y_true = sv1_df$Ecoli_Status, positive = "Ecoli_LOS")

# F1 Score
F1_val <- F1_Score(y_pred = sv1_class, y_true = sv1_df$Ecoli_Status, positive = "Ecoli_LOS")

# Print
cat("Precision:", Precision_val, "\n")
cat("F1 Score:", F1_val, "\n")

```






```{r}
sum(sv1_class == "Ecoli_LOS")  # how many predicted positives?
sum(sv1_df$Ecoli_Status == "Ecoli_LOS")  # how many true positives?
```



```{r}
# --- SV2 Metrics ---
sv2_cm <- confusionMatrix(sv2_class, sv2_df$Ecoli_Status, positive = "Ecoli_LOS")

# View
sv2_cm
```


```{r}
# Predictions

#old
# ------------------------------------------------------------
sv1_class <- predict(rf_model, sv1_df)
sv1_prob  <- predict(rf_model, sv1_df, type = "prob")[, "Ecoli_LOS"]

# ------------------------------------------------------------
# Confusion matrix
# ------------------------------------------------------------
cm <- confusionMatrix(sv1_class, sv1_df$Ecoli_Status, positive = "Ecoli_LOS")

# ------------------------------------------------------------
# Extract metrics
# ------------------------------------------------------------
Precision   <- Precision(y_pred = sv1_class, y_true = sv1_df$Ecoli_Status, positive = "Ecoli_LOS")
F1_score    <- F1_Score(y_pred = sv1_class, y_true = sv1_df$Ecoli_Status, positive = "Ecoli_LOS")
Sensitivity <- cm$byClass["Sensitivity"]
Specificity <- cm$byClass["Specificity"]

# AUROC
roc_obj <- roc(response = sv1_df$Ecoli_Status,
               predictor = sv1_prob,
               levels = c("Control", "Ecoli_LOS"),
               direction = "<")
AUROC <- auc(roc_obj)

# ------------------------------------------------------------
# Combine into a single-row table
# ------------------------------------------------------------
summary_metrics <- data.frame(
  Precision   = Precision,
  F1_score    = F1_score,
  Sensitivity = Sensitivity,
  Specificity = Specificity,
  AUROC       = as.numeric(AUROC)
)

print(summary_metrics)

```


```{r}

# ------------------------------------------------------------
# Predictions
# ------------------------------------------------------------
sv2_class <- predict(rf_model, sv2_df)
sv2_prob  <- predict(rf_model, sv2_df, type = "prob")[, "Ecoli_LOS"]

# ------------------------------------------------------------
# Confusion matrix
# ------------------------------------------------------------
cm <- confusionMatrix(sv2_class, sv2_df$Ecoli_Status, positive = "Ecoli_LOS")

# ------------------------------------------------------------
# Extract metrics
# ------------------------------------------------------------
Precision   <- Precision(y_pred = sv2_class, y_true = sv2_df$Ecoli_Status, positive = "Ecoli_LOS")
F1_score    <- F1_Score(y_pred = sv2_class, y_true = sv2_df$Ecoli_Status, positive = "Ecoli_LOS")
Sensitivity <- cm$byClass["Sensitivity"]
Specificity <- cm$byClass["Specificity"]

# AUROC
roc_obj <- roc(response = sv2_df$Ecoli_Status,
               predictor = sv2_prob,
               levels = c("Control", "Ecoli_LOS"),
               direction = "<")
AUROC <- auc(roc_obj)

# ------------------------------------------------------------
# Combine into a single-row table
# ------------------------------------------------------------
summary_metrics <- data.frame(
  Precision   = Precision,
  F1_score    = F1_score,
  Sensitivity = Sensitivity,
  Specificity = Specificity,
  AUROC       = as.numeric(AUROC)
)

print(summary_metrics)

```



```{r}


```

