
# 

#Load libraries and data:

Required libraries:
```{r}
library(vegan)
library (ggplot2)
library(data.table)
library(cowplot)
library(Maaslin2)
library(readr)
library(phyloseq)
library(pgirmess)
library(agricolae)
library(readxl) 
library(caret)
library(tidyverse)
library(microbiome)
library(MLmetrics)
library(randomForest)
library(gridExtra)
library(grid)
library(pROC)  # for ROC/AUC

```

#How to make phyloseq object:

Reformat table:


```{r}
# ------------------------------------------------------------
# 2️⃣ Read OTU table (Discovery cohort - DC)
# ------------------------------------------------------------
discDC <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                     sheet = "DC") %>% as.data.frame()

# First column contains sample IDs → make them rownames
rownames(discDC) <- discDC[[1]]
discDC <- discDC[ , -1]

# Remove unwanted PC samples (rows)
remove_samples <- paste0("DC-PC", 1:13)
discDC <- discDC[!rownames(discDC) %in% remove_samples, ]
cat("Number of samples after removing PC samples:", nrow(discDC), "\n")

# Transpose: rows = OTUs, cols = samples
DCotumat <- t(discDC)

# ------------------------------------------------------------
# 3️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# Keep only OTUs present in both OTU and taxonomy tables
common_otus <- intersect(rownames(DCotumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# Subset OTU and taxonomy tables
DCotumat_common <- DCotumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# Sanity check
dim(DCotumat_common)
dim(taxonomy_common)

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(DCotumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqDC <- phyloseq(OTU, TAX)
cat("physeqDC - Samples:", nsamples(physeqDC), "OTUs:", ntaxa(physeqDC), "\n")

## new added
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Discovery cohort", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"

# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(DCotumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqDC <- merge_phyloseq(physeqDC, sampledata)

cat("physeqDC with metadata - Samples:", nsamples(physeqDC), 
    "OTUs:", ntaxa(physeqDC), "\n")


```


# If you also have SV1 and SV2 phyloseq objects

# prep SV1
```{r}

# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV1
# ------------------------------------------------------------
discSV1 <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx", 
                      sheet = "SV 1") %>% as.data.frame()

# first column = sample IDs
rownames(discSV1) <- discSV1[[1]]
discSV1 <- discSV1[ , -1]

# transpose: rows = OTUs, cols = samples
SV1otumat <- t(discSV1)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV1otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

# subset OTU and taxonomy tables
SV1otumat_common <- SV1otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV1otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV1 <- phyloseq(OTU, TAX)

cat("physeqSV1 - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 1
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 1", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"
# Keep only rows that exist in SV1 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV1otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV1 <- merge_phyloseq(physeqSV1, sampledata)

cat("physeqSV1 with metadata - Samples:", nsamples(physeqSV1), 
    "OTUs:", ntaxa(physeqSV1), "\n")

```

# prep SV2 data
```{r}
# ------------------------------------------------------------
# 1️⃣ Read OTU table - SV2
# ------------------------------------------------------------
discSV2 <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx", 
                      sheet = "SV 2") %>% as.data.frame()

# Suppose the first column contains the row IDs (sample names)
rownames(discSV2) <- discSV2[[1]]
discSV2 <- discSV2[ , -1]  # remove the first column after assigning rownames

# Now remove rows that contain "SV2-Blanco"
discSV2 <- discSV2[!grepl("^SV2-Blanco", rownames(discSV2)), ]

cat("Number of rows after removing blanks:", nrow(discSV2), "\n")

# transpose: rows = OTUs, cols = samples
SV2otumat <- t(discSV2)

# ------------------------------------------------------------
# 2️⃣ Read taxonomy table
# ------------------------------------------------------------
taxonomy <- read_excel("20250512_OTU tables_percohort_sepsis prediction model.xlsx",
                       sheet = "OTU table all") %>% as.data.frame()

# Use correct OTU ID column (ID.1) as rownames
rownames(taxonomy) <- taxonomy$ID.1
taxonomy <- taxonomy[ , !(colnames(taxonomy) %in% "ID.1") ]

# ------------------------------------------------------------
# 3️⃣ Keep only OTUs present in both OTU and taxonomy tables
# ------------------------------------------------------------
common_otus <- intersect(rownames(SV2otumat), rownames(taxonomy))
cat("Number of OTUs in both OTU and taxonomy tables:", length(common_otus), "\n")

SV2otumat_common <- SV2otumat[common_otus, , drop=FALSE]
taxonomy_common <- taxonomy[common_otus, , drop=FALSE]

# ------------------------------------------------------------
# 4️⃣ Build phyloseq object
# ------------------------------------------------------------
OTU <- otu_table(as.matrix(SV2otumat_common), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy_common))
physeqSV2 <- phyloseq(OTU, TAX)

cat("physeqSV2 - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

# ------------------------------------------------------------
# 5️⃣ Add metadata
# ------------------------------------------------------------
meta_data <- read_excel("20250611_Overview samples for Animesh.xlsx", 
                        sheet = "Metadata") %>% as.data.frame()

# Filter for validation cohort 2
meta_data <- meta_data[meta_data$Cohort == "Validatie cohort 2", ]

# Rename columns
colnames(meta_data)[7] <- "Timeperiods"
colnames(meta_data)[9] <- "DiseaseStatus"
colnames(meta_data)[10] <- "DiseaseCausative"
#colnames(meta_data)[11] <- "Bloodculture_isolated_pathogen"
# Keep only rows that exist in SV2 OTU table
meta_data_filtered <- meta_data[meta_data$Sample_ID %in% colnames(SV2otumat_common), ]

# Set rownames to match sample names in physeq
rownames(meta_data_filtered) <- meta_data_filtered$Sample_ID

# Create sample_data object
sampledata <- sample_data(meta_data_filtered)

# Merge with phyloseq
physeqSV2 <- merge_phyloseq(physeqSV2, sampledata)

cat("physeqSV2 with metadata - Samples:", nsamples(physeqSV2), 
    "OTUs:", ntaxa(physeqSV2), "\n")

```

# til sv2 prep 

## newruns for Model for E.coli sepsis with overlapping ASVX 

```{r}
# ------------------------------------------------------------
# Compute common OTUs across Discovery (DC), VC1, VC2 cohorts
# ------------------------------------------------------------

# Extract OTU tables (no species collapsing)
otu_full <- as.data.frame(t(otu_table(physeqDC)))    # Discovery cohort
otu_sv1  <- as.data.frame(t(otu_table(physeqSV1)))   # Validation 1
otu_sv2  <- as.data.frame(t(otu_table(physeqSV2)))   # Validation 2

# Find OTUs present in all three cohorts
common_otus <- Reduce(intersect, list(
  colnames(otu_full),
  colnames(otu_sv1),
  colnames(otu_sv2)
))

cat("Number of common OTUs:", length(common_otus), "\n")


```

# E. coli LOS vs Control (Discovery Cohort, all time periods)


```{r}

# ------------------------------------------------------------
# 2️⃣ Prepare metadata for E. coli LOS vs Control comparison
# ------------------------------------------------------------
meta_ecoli <- data.frame(sample_data(physeqDC))

# Recode DiseaseStatus numeric to text
meta_ecoli$DiseaseStatus_txt <- ifelse(meta_ecoli$DiseaseStatus == 1, "LOS", "Control")

# Define Ecoli_Status
meta_ecoli$Ecoli_Status <- NA

# Label E. coli LOS cases
meta_ecoli$Ecoli_Status[
  meta_ecoli$Bloodculture_isolated_pathogen == "Escherichia coli" &
  meta_ecoli$DiseaseStatus_txt == "LOS"
] <- "Ecoli_LOS"

# Label Controls (all controls, regardless of pathogen)
meta_ecoli$Ecoli_Status[
  meta_ecoli$DiseaseStatus_txt == "Control"
] <- "Control"

# Check distribution
table(meta_ecoli$Ecoli_Status)


# Subset phyloseq to only these samples
physeq_ecoli <- prune_samples(rownames(meta_ecoli), physeqDC)


# ------------------------------------------------------------
# 3️⃣ Extract OTU table (common OTUs, not species)
# ------------------------------------------------------------
otu_ecoli <- as.data.frame(t(otu_table(physeq_ecoli)))

# Filter to only common OTUs used across cohorts
otu_ecoli <- otu_ecoli[, common_otus, drop = FALSE]
  # reuse from previous step (common_species == common OTUs)

# Add pseudocount and apply CLR + Z-normalization
otu_ecoli_pseudo <- otu_ecoli + 1
otu_ecoli_clr <- microbiome::transform(otu_ecoli_pseudo, "clr")
otu_ecoli_clr_z <- scale(otu_ecoli_clr)

# Save scaling parameters for validation
train_mean <- attr(otu_ecoli_clr_z, "scaled:center")
train_sd   <- attr(otu_ecoli_clr_z, "scaled:scale")


# ------------------------------------------------------------
# 4️⃣ Combine with metadata
# ------------------------------------------------------------
df_ecoli <- as.data.frame(otu_ecoli_clr_z)
df_ecoli$Ecoli_Status <- factor(meta_ecoli$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

cat("Number of predictors:", ncol(df_ecoli) - 1, "\n")
cat("Samples per class:\n")
print(table(df_ecoli$Ecoli_Status))

```

```{r}

# ------------------------------------------------------------
# 1️⃣ Ensure no missing values
# ------------------------------------------------------------
# Remove samples with NA in Ecoli_Status
# Keep only samples with known Ecoli_Status
df_ecoli_clean <- df_ecoli[!is.na(df_ecoli$Ecoli_Status), ]

# Ensure factor levels are correct
df_ecoli_clean$Ecoli_Status <- factor(df_ecoli_clean$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

# Optional: replace any remaining NA in OTUs
df_ecoli_clean[is.na(df_ecoli_clean)] <- 0

# Check distribution
table(df_ecoli_clean$Ecoli_Status)

# 2️⃣ Split into training/testing
set.seed(123)
train_index <- createDataPartition(df_ecoli_clean$Ecoli_Status, p = 0.8, list = FALSE)
train_set   <- df_ecoli_clean[train_index, ]
test_set    <- df_ecoli_clean[-train_index, ]

# ------------------------------------------------------------
# 3️⃣ Custom summary function for caret
# ------------------------------------------------------------
multiSummary <- function(data, lev = NULL, model = NULL) {
  roc_val <- tryCatch({
    roc(response = data$obs, predictor = data[, lev[2]], levels = rev(lev), quiet = TRUE)$auc
  }, error = function(e) NA)
  
  out <- c(
    ROC       = roc_val,
    Sens      = sensitivity(data$pred, data$obs, positive = lev[2]),
    Spec      = specificity(data$pred, data$obs, negative = lev[1]),
    Precision = Precision(y_pred = data$pred, y_true = data$obs, positive = lev[2]),
    Recall    = Recall(y_pred = data$pred, y_true = data$obs, positive = lev[2]),
    F1        = F1_Score(y_pred = data$pred, y_true = data$obs, positive = lev[2])
  )
  return(out)
}


# ------------------------------------------------------------
# Train Random Forest (ranger)
# ------------------------------------------------------------
set.seed(123)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 3,     # smaller folds for rare class else use 10
  repeats = 5,
  classProbs = TRUE,
  summaryFunction = multiSummary,
  sampling = "up",   # upsample minority class
  savePredictions = "final"
)

#  Check class distribution
table(train_set$Ecoli_Status)
table(test_set$Ecoli_Status)

rf_model <- train(
  Ecoli_Status ~ .,
  data       = df_ecoli_clean,
  method     = "ranger",
  metric     = "F1",          # or "Accuracy"
  tuneGrid   = expand.grid(
                 mtry          = floor(sqrt(ncol(df_ecoli_clean) - 1)),
                 splitrule     = "gini",
                 min.node.size = 1
               ),
  trControl  = ctrl,
  importance = "impurity"
)


# ------------------------------------------------------------
# Extract and summarize CV results
# ------------------------------------------------------------
results <- rf_model$resample[, c("ROC","Sens","Spec","Precision","Recall","F1")]

summary_table <- data.frame(
  Average = colMeans(results, na.rm = TRUE),
  SD      = apply(results, 2, sd, na.rm = TRUE)
)

summary_table <- t(summary_table)
print(summary_table)

```

```{r}
# ------------------------------------------------------------
# 6️⃣ Extract and summarize CV results
# ------------------------------------------------------------

# Use the model object returned from train()
results <- rf_model$resample[, c("ROC","Sens","Spec","Precision","Recall","F1")]

# Check if results has at least 1 row
if(nrow(results) == 0){
  stop("No resample results found. Check if training produced valid output.")
}

# Compute average and SD
summary_table <- data.frame(
  Average = colMeans(results, na.rm = TRUE),
  SD      = apply(results, 2, sd, na.rm = TRUE)
)

# Transpose so rows = Average/SD, columns = metrics
summary_table <- t(summary_table)

print(summary_table)

```


###  end this DC ML ###





# only for DC - top 20 otus

```{r}

# Extract variable importance
var_imp <- varImp(rf_model)$importance

# Add Original_ID column directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending
var_imp_ordered <- var_imp[order(-var_imp$Overall), ]

# Select top 20
top20 <- head(var_imp_ordered, 20)

# Plot
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()


```


# change every time file name 
```{r}

# --- 1️⃣ Extract variable importance ---
var_imp <- varImp(rf_model)$importance

# Add Original_ID directly from rownames
var_imp$Original_ID <- rownames(var_imp)

# Order by importance descending and add Rank
var_imp <- var_imp[order(-var_imp$Overall), ]
var_imp$Rank <- 1:nrow(var_imp)

# Top 20
top20 <- head(var_imp, 20)

# --- 2️⃣ Plot top 20 OTUs ---
ggplot(top20, aes(x = reorder(Original_ID, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Original-ID", y = "Importance", title = "Top 20 OTUs by Importance") +
  theme_minimal()

# --- 3️⃣ Merge taxonomy info ---
tax <- read_excel(
  "20250512_OTU tables_percohort_sepsis prediction model.xlsx",
  sheet = "OTU table all"
) %>%
  mutate(ID.1 = as.character(ID.1))

top20 <- top20 %>%
  mutate(Original_ID = as.character(Original_ID)) %>%
  left_join(tax %>% select(ID.1, Family, Genus, Species),
            by = c("Original_ID" = "ID.1")) %>%
  select(Rank, Original_ID, Family, Genus, Species, Overall)

# --- 4️⃣ Create table grob ---
tab_grob <- tableGrob(
  top20,
  rows = NULL,
  theme = ttheme_default(
    core   = list(fg_params = list(cex = 0.8)),
    colhead= list(fg_params = list(cex = 1, fontface = "bold"))
  )
)

title <- textGrob("Top 20 OTUs with Taxonomy",
                  gp = gpar(fontsize = 16, fontface = "bold"))

# --- 5️⃣ Save as PNG ---
png("ecoli_Top20_OTU_Taxonomy.png", width = 2000, height = 1400, res = 200)
grid.arrange(title, tab_grob, ncol = 1, heights = c(0.1, 0.8, 0.1))
dev.off()

```



# now test in SVs

# --- Function to prepare validation set ---
```{r}
# --- Function to prepare validation set ---

prepare_validation <- function(physeq_obj, train_mean, train_sd, dc_features) {
  
  # 1️⃣ Extract OTU table and transpose
  otu_df <- as.data.frame(t(otu_table(physeq_obj)))
  
  # 2️⃣ Remove duplicate OTU names (if any)
  otu_df <- otu_df[, !duplicated(colnames(otu_df)), drop = FALSE]
  
  # 3️⃣ Add missing OTUs as zeros
  missing_feats <- setdiff(dc_features, colnames(otu_df))
  if(length(missing_feats) > 0) {
    warning("Adding missing OTUs as zeros: ", paste(missing_feats, collapse = ", "))
    otu_df[, missing_feats] <- 0
  }
  
  # 4️⃣ Keep only DC OTUs and reorder exactly
  otu_df <- otu_df[, dc_features, drop = FALSE]
  
  # 5️⃣ Add pseudocount
  otu_df <- otu_df + 1
  
  # 6️⃣ CLR transform
  log_mat <- log(otu_df)
  row_means <- rowMeans(log_mat)
  otu_clr <- sweep(log_mat, 1, row_means, "-")
  
  # 7️⃣ Check dimensions before scaling
  if(ncol(otu_clr) != length(train_mean) || ncol(otu_clr) != length(train_sd)) {
    stop(
      "Mismatch in number of columns: ",
      "OTU table has ", ncol(otu_clr), 
      ", train_mean has ", length(train_mean),
      ", train_sd has ", length(train_sd)
    )
  }
  
  # 8️⃣ Z-normalize
  otu_clr_z <- scale(otu_clr, center = train_mean, scale = train_sd)
  
  # 9️⃣ Metadata & labels
  meta <- data.frame(sample_data(physeq_obj))
  meta$Ecoli_Status <- NA
  meta$Ecoli_Status[
    meta$Bloodculture_isolated_pathogen == "Escherichia coli" &
    meta$DiseaseStatus %in% c(1, "LOS")
  ] <- "Ecoli_LOS"
  meta$Ecoli_Status[
    meta$DiseaseStatus %in% c(0, "Control")
  ] <- "Control"
  
  meta$Ecoli_Status <- factor(meta$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))
  
  # 🔟 Combine OTU + labels
  df <- as.data.frame(otu_clr_z)
  df$Ecoli_Status <- meta$Ecoli_Status
  df <- df[!is.na(df$Ecoli_Status), ]
  
  return(df)
}


# Use only OTU columns, exclude Ecoli_Status
dc_features <- setdiff(colnames(df_ecoli), "Ecoli_Status")

sv1_df <- prepare_validation(physeqSV1, train_mean, train_sd, dc_features)
sv2_df <- prepare_validation(physeqSV2, train_mean, train_sd, dc_features)

nrow(sv1_df); nrow(sv2_df)
```
## 
```{r}
# Predictions on SV1
sv1_pred <- predict(rf_model, sv1_df, type = "prob")
sv1_class <- predict(rf_model, sv1_df)

# Predictions on SV2
sv2_pred <- predict(rf_model, sv2_df, type = "prob")
sv2_class <- predict(rf_model, sv2_df)

# --- Function to compute ROC/AUC ---
# SV1 ROC/AUC
compute_roc <- function(model, df, positive_class = "Ecoli_LOS") {
  probs <- predict(model, df, type = "prob")[, positive_class]
  truth <- df$Ecoli_Status  # make sure to use Ecoli_Status, not DiseaseStatus
  roc_obj <- roc(response = truth,
                 predictor = probs,
                 levels = c("Control", positive_class),
                 direction = "<")
  auc_val <- auc(roc_obj)
  plot(roc_obj, main = paste0("ROC curve (AUC = ", round(auc_val, 3), ")"))
  return(list(roc = roc_obj, auc = auc_val))
}

# SV1
sv1_roc <- compute_roc(rf_model, sv1_df)

# SV2
sv2_roc <- compute_roc(rf_model, sv2_df)



```







```{r}
## Make sure the factor levels match
sv1_df$Ecoli_Status <- factor(sv1_df$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))
sv2_df$Ecoli_Status <- factor(sv2_df$Ecoli_Status, levels = c("Control", "Ecoli_LOS"))

# Predictions
sv1_class <- predict(rf_model, sv1_df)
sv2_class <- predict(rf_model, sv2_df)

# Confusion matrices
library(caret)
sv1_cm <- confusionMatrix(sv1_class, sv1_df$Ecoli_Status, positive = "Ecoli_LOS")#
sv1_cm


```
```{r}
# --- SV2 Metrics ---
sv2_cm <- confusionMatrix(sv2_class, sv2_df$Ecoli_Status, positive = "Ecoli_LOS")

# View
sv2_cm
```


```{r}
library(caret)
library(pROC)
library(MLmetrics)

# Predicted classes and probabilities
sv1_class <- predict(rf_model, sv1_df)
sv1_prob  <- predict(rf_model, sv1_df, type = "prob")[, "Ecoli_LOS"]  # probability for LOS

# Confusion Matrix
cm <- confusionMatrix(sv1_class, sv1_df$DiseaseStatus, positive = "Ecoli_LOS")

# Extract Accuracy, Sensitivity, Specificity
accuracy    <- cm$overall["Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]

# Precision
precision <- Precision(y_pred = sv1_class, y_true = sv1_df$DiseaseStatus, positive = "LOS")

# F1 Score
f1 <- F1_Score(y_pred = sv1_class, y_true = sv1_df$DiseaseStatus, positive = "LOS")

# AUC
roc_obj <- roc(response = sv1_df$DiseaseStatus,
               predictor = sv1_prob,
               levels = c("Control","LOS"), direction = "<")
auc_val <- auc(roc_obj)

# Print all
metrics <- data.frame(
  Accuracy    = accuracy,
  Sensitivity = sensitivity,
  Specificity = specificity,
  Precision   = precision,
  F1          = f1,
  AUC         = as.numeric(auc_val)
)
print(metrics)

```


```{r}
library(caret)
library(pROC)
library(MLmetrics)

# Predicted classes and probabilities
sv2_class <- predict(rf_model, sv2_df)
sv2_prob  <- predict(rf_model, sv2_df, type = "prob")[, "LOS"]  # probability for LOS

# Confusion Matrix
cm <- confusionMatrix(sv2_class, sv2_df$DiseaseStatus, positive = "LOS")

# Extract Accuracy, Sensitivity, Specificity
accuracy    <- cm$overall["Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]

# Precision
precision <- Precision(y_pred = sv2_class, y_true = sv2_df$DiseaseStatus, positive = "LOS")

# F1 Score
f1 <- F1_Score(y_pred = sv2_class, y_true = sv2_df$DiseaseStatus, positive = "LOS")

# AUC
roc_obj <- roc(response = sv2_df$DiseaseStatus,
               predictor = sv2_prob,
               levels = c("Control","LOS"), direction = "<")
auc_val <- auc(roc_obj)

# Print all
metrics <- data.frame(
  Accuracy    = accuracy,
  Sensitivity = sensitivity,
  Specificity = specificity,
  Precision   = precision,
  F1          = f1,
  AUC         = as.numeric(auc_val)
)
print(metrics)

```
```{r}
identical(sample_names(physeqSV1), sample_names(physeqSV2))
nrow(sv1_df); nrow(sv2_df)

```

